---
title: "Evaluate drug stats"
author: "Sara gosline"
date: "2025-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(synapser)
library(ggplot2)
library(dplyr)
library(tidyr)


```

## Download drug stats
First we need to downlod the summary stats for each organoid 
```{r download, warning=FALSE}
synLogin()

fits<-readr::read_tsv(synGet('syn65471817')$path)


meta<-readxl::read_xlsx(syn$get('syn65595365')$path)|>
  tidyr::separate(Specimen,into=c('Patient','Tumor'),sep='_',remove = FALSE)

```

## Plot data from drug

Here we can write a function that plots data for a drug across patients

```{r summary, echo=FALSE}

 fits|>subset(dose_response_metric=='uM_viability')|>ggplot(aes(x=dose_response_value,y=improve_sample_id,fill=improve_sample_id,alpha=0.5))+ggridges::geom_density_ridges()+ggtitle('Viabilities across samples')

##we can also plot samples according to their drug response
pcs = fits|>subset(dose_response_metric=='uM_viability')|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fill=1.0)|>
  tibble::column_to_rownames('improve_sample_id')|>prcomp()

pcs$x|>as.data.frame()|>tibble::rownames_to_column('Specimen')|>
  left_join(meta)|>
  dplyr::select(PC1,PC2,Patient,Specimen)|>
  distinct()|>
  ggplot(aes(x=PC1,y=PC2,col=Patient,label=Specimen))+geom_point()+ggrepel::geom_label_repel()

```

## Now compare to proteomics

We created a metadata file with the aliquot number that is embedded in the file name/columns, and teased it out to create two long tables - one for global and one for phospho.

```{r compare to proteomics,warning=FALSE}




phospho<- read.table(syn$get('syn65467785')$path)|>
  tibble::rownames_to_column('site')

##logtransform##median transform

pmat<-apply(log2(phospho[,2:ncol(phospho)]),2,function(x) x-median(x,na.rm=T))|>
  as.data.frame()|>
  mutate(site=phospho$site)

plong<-pmat|>
  tidyr::pivot_longer(1:(ncol(pmat)-1),names_to='fname',values_to='abundance')|>
  mutate(aliquot=sapply(fname,function(x) unlist(strsplit(x,split='_'))[9]))|>
  mutate(aliquot=as.double(aliquot))|>
  left_join(meta)

##save to file
readr::write_csv(plong,file='log2normMedCenteredPhospho.csv')
syn$store(File('log2normMedCenteredPhospho.csv',parentId='syn51301417'))
         

plong$abundance[which(!is.finite(plong$abundance))]<-0
ppcs<-plong|>dplyr::select(Specimen,abundance,site)|>
  subset(!is.na(site))|>
  subset(!is.na(abundance))|>
  tidyr::pivot_wider(names_from='Specimen',values_from='abundance',values_fn=mean,values_fill=0)|>
  tibble::column_to_rownames('site')|>t()|>
  prcomp()

pplot<-ppcs$x|>
  as.data.frame()|>
  dplyr::select(PC1,PC2)|>
  tibble::rownames_to_column('Specimen')|>
  left_join(meta)|>
  dplyr::select(PC1,PC2,Specimen,Patient)|>
  distinct()|>
  ggplot(aes(x=PC1,y=PC2,label=Specimen,col=Patient))+geom_point()+ggrepel::geom_label_repel()+ggtitle("Phospho samples")

pplot

global<-readr::read_tsv(syn$get('syn64906445')$path)

##logtransform, medina transform

gmat<-apply(log2(global[,6:ncol(global)]),2,function(x) x-median(x,na.rm=T))

glong<-gmat|>
  as.data.frame()|>
  mutate(Genes=global$Genes)|>
  tidyr::pivot_longer(1:ncol(gmat),names_to='fname',values_to='abundance')|>
  mutate(aliquot=sapply(fname,function(x) unlist(strsplit(x,split='_'))[9]))|>
  mutate(aliquot=as.double(aliquot))|>
  left_join(meta)
         

readr::write_csv(plong,file='log2normMedCenteredGlobal.csv')
syn$store(File('log2normMedCenteredGlobal.csv',parentId='syn51301417'))
         
#ma<-mean(glong$abundance,na.rm=T)
glong$abundance[which(!is.finite(glong$abundance))]<-0

gpcs<-glong|>dplyr::select(Specimen,abundance,Genes)|>
  subset(!is.na(Genes))|>
  subset(!is.na(abundance))|>
  tidyr::pivot_wider(names_from='Specimen',values_from='abundance',values_fn=mean,values_fill=0)|>
  tibble::column_to_rownames('Genes')|>t()|>
  prcomp()

gplot<-gpcs$x|>
  as.data.frame()|>
  dplyr::select(PC1,PC2)|>
  tibble::rownames_to_column('Specimen')|>
  left_join(meta)|>
    dplyr::select(PC1,PC2,Specimen,Patient)|>
  distinct()|>
  ggplot(aes(x=PC1,y=PC2,label=Specimen,col=Patient))+geom_point()+ggrepel::geom_label_repel()+ggtitle("Global samples")

gplot

```
Now we have the global and phospho and drug data for anumber of samples. 

## Format protein data to collect correlation values

Do simple correlations to identify putative trends in the data.

```{r make matrices of proteins}

shared<-intersect(fits$improve_sample_id,glong$Specimen)
print(paste('Found',length(shared),'shared samples'))

glob_dat<-glong|>
  subset(Specimen%in%shared)|>
  dplyr::select(Specimen,Genes,abundance)|>
  tidyr::pivot_wider(names_from='Genes',values_from='abundance',values_fill=0,values_fn=mean)|>
  tibble::column_to_rownames('Specimen')
  

phos_dat<-plong|>
  subset(Specimen%in%shared)|>
  dplyr::select(Specimen,site,abundance)|>
  tidyr::pivot_wider(names_from='site',values_from='abundance',values_fill=0,values_fn=mean)|>
  tibble::column_to_rownames('Specimen')


```

### Correlate uM viabilities with protein/phosphosite expression

Look at uM viabilities and see what we can find that correlates with expression

```{r find common samples, warning=FALSE,error=FALSE}


## a full join might be a challenge, maybe just take two matrices
drug_dat <- fits|>subset(dose_response_metric=='uM_viability')|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fn=mean)|>
  tibble::column_to_rownames('improve_sample_id')

gres<-cor(drug_dat[shared,],glob_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
  as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(glob_dat)),names_to='gene',values_to='pCor')|>
  arrange(desc(pCor))

##now lets try to get significance

gsig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(glob_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],glob_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>
  as.data.frame()|>
  mutate(drug=unlist(drug))|>
  mutate(gene=unlist(gene))

fullcors<-gres|>left_join(data.frame(gsig))|>mutate(data='global')

pres<-cor(drug_dat[shared,],phos_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
    as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(phos_dat)),names_to='site',values_to='pCor')|>
  arrange(desc(pCor))

##now lets look at correlations

psig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(phos_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],phos_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>as.data.frame()

fullpcors<-pres|>rename(gene='site')|>left_join(data.frame(psig))|>mutate(data='phospho')

#combine all correlations
allcor<-rbind(fullcors,fullpcors)|>
  mutate(direction=ifelse(pCor<0,'neg','pos'))


##lets count the correlations and plot

corsummary<-allcor|>subset(bh_p<0.1)|>
  group_by(drug,data,direction)|>
  summarize(features=n(),meanCor=mean(pCor))
corsummary|>
  subset(features>1)|>
  ggplot(aes(x=data,y=features,fill=drug))+facet_grid(~direction)+geom_bar(position='dodge',stat='identity')

```
Now that we have the correlations we can cherry pick some examples to showcase

```{r plot cors}

plotCors<-function(featureslist,druglist,dataType='global'){
  ##subset a list of features and drugs and plot those in a graph
  require(ggplot2)
  if(dataType=='global'){
    ptab<-glong|>dplyr::rename(feature='Genes')
  }else{
    ptab<-plong|>dplyr::rename(feature='site')
  }
  dtab<-fits|>
    subset(dose_response_metric=='uM_viability')|>
    dplyr::rename(Specimen='improve_sample_id',Drug='improve_drug_id')|>
    subset(Drug%in%druglist)
  
  ftab<-subset(ptab,feature%in%featureslist)|>left_join(dtab)|>
    subset(!is.na(Drug))
  ftab|>ggplot(aes(x=abundance,y=dose_response_value,col=Drug,shape=Patient))+geom_point()+facet_grid(feature~.)
}

druglist<-c('Onalespib','Crizotinib')
features<-subset(allcor,drug%in%druglist)|>
  subset(bh_p<0.05)|>
  subset(data=='global')

plotCors(features$gene,druglist)


```

now we can visualize correlations
```{r check out HSP90s}

hsps<-unique(glong$Genes[grep('^HSP',glong$Genes)])

cor_hsps<-subset(allcor,gene%in%hsps)|>subset(drug=='Onalespib')|>subset(corp<0.05)
print(paste('measured',length(hsps),'HSPs in global data of which', nrow(cor_hsps),' are correlated with Onalespib'))

plotCors(cor_hsps$gene,'Onalespib')

hspp<-unique(plong$site[grep('^HSP',plong$site)])
cor_hspps<-subset(allcor,gene%in%hspp)|>subset(drug=='Onalespib')|>subset(corp<0.05)
print(paste('measured',length(hspp),'HSPs in phospho data of which',nrow(cor_hspps),' are correlated with Onalespib'))


```

### Now lets look only at IC50 values

There are a few drugs for which we have IC50 values

```{r check ic50 cors,warning=FALSE,error=FALSE}

ifits<-subset(fits,dose_response_metric=='fit_ic50')

shared<-intersect(ifits$improve_sample_id,glong$Specimen)
print(paste('Found',length(shared),'shared samples'))

## a full join might be a challenge, maybe just take two matrices
drug_dat <- ifits|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fn=mean)|>
  tibble::column_to_rownames('improve_sample_id')

gres<-cor(drug_dat[shared,],glob_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
  as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(glob_dat)),names_to='gene',values_to='pCor')|>
  arrange(desc(pCor))

##now lets try to get significance

gsig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(glob_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],glob_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>
  as.data.frame()|>
  mutate(drug=unlist(drug))|>
  mutate(gene=unlist(gene))

fullcors<-gres|>left_join(data.frame(gsig))|>mutate(data='global')

pres<-cor(drug_dat[shared,],phos_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
    as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(phos_dat)),names_to='site',values_to='pCor')|>
  arrange(desc(pCor))

##now lets look at correlations

psig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(phos_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],phos_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>as.data.frame()

fullpcors<-pres|>rename(gene='site')|>left_join(data.frame(psig))|>mutate(data='phospho')

#combine all correlations
allcor<-rbind(fullcors,fullpcors)|>
  mutate(direction=ifelse(pCor<0,'neg','pos'))


##lets count the correlations and plot

corsummary<-allcor|>subset(bh_p<0.1)|>
  group_by(drug,data,direction)|>
  summarize(features=n(),meanCor=mean(pCor))

corsummary|>
  #subset(features>1)|>
  ggplot(aes(x=data,y=features,fill=drug))+facet_grid(~direction)+geom_bar(position='dodge',stat='identity')

```

Again we have onalespib with numerous significantly correlated proteins, and one phosphosite for digoxin showing up .

```{r plot individual sites}

druglist<-c('Onalespib')
features<-subset(allcor,drug%in%druglist)|>
  subset(bh_p<0.1)|>
  subset(data=='global')

plotCors(features$gene,druglist)


druglist<-c('Digoxin')
features<-subset(allcor,drug%in%druglist)|>
  subset(bh_p<0.1)|>
  subset(data=='phospho')

plotCors(features$gene,druglist,data='phospho')

```

Now we can check the HSP proteins directly

```{r HSP correlation}
hsps<-unique(glong$Genes[grep('^HSP',glong$Genes)])

cor_hsps<-subset(allcor,gene%in%hsps)|>subset(drug=='Onalespib')|>subset(corp<0.05)
print(paste('measured',length(hsps),'HSPs in global data of which', nrow(cor_hsps),' are correlated with Onalespib'))

plotCors(cor_hsps$gene,'Onalespib')

hspp<-unique(plong$site[grep('^HSP',plong$site)])
cor_hspps<-subset(allcor,gene%in%hspp)|>subset(drug=='Onalespib')|>subset(corp<0.1)
print(paste('measured',length(hspp),'HSPs in phospho data of which',nrow(cor_hspps),' are correlated with Onalespib'))
```

The IC50 result is similar to the viability. Now we can check AUC

```{r auc hsp check}

ifits<-subset(fits,dose_response_metric=='fit_auc')

shared<-intersect(ifits$improve_sample_id,glong$Specimen)
print(paste('Found',length(shared),'shared samples'))

## a full join might be a challenge, maybe just take two matrices
drug_dat <- ifits|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fn=mean)|>
  tibble::column_to_rownames('improve_sample_id')

gres<-cor(drug_dat[shared,],glob_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
  as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(glob_dat)),names_to='gene',values_to='pCor')|>
  arrange(desc(pCor))

##now lets try to get significance

gsig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(glob_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],glob_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>
  as.data.frame()|>
  mutate(drug=unlist(drug))|>
  mutate(gene=unlist(gene))

fullcors<-gres|>left_join(data.frame(gsig))|>mutate(data='global')

pres<-cor(drug_dat[shared,],phos_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
    as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(phos_dat)),names_to='site',values_to='pCor')|>
  arrange(desc(pCor))

##now lets look at correlations

psig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(phos_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],phos_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>as.data.frame()

fullpcors<-pres|>rename(gene='site')|>left_join(data.frame(psig))|>mutate(data='phospho')

#combine all correlations
allcor<-rbind(fullcors,fullpcors)|>
  mutate(direction=ifelse(pCor<0,'neg','pos'))


##lets count the correlations and plot

corsummary<-allcor|>subset(bh_p<0.1)|>
  group_by(drug,data,direction)|>
  summarize(features=n(),meanCor=mean(pCor))

corsummary|>
  #subset(features>1)|>
  ggplot(aes(x=data,y=features,fill=drug))+facet_grid(~direction)+geom_bar(position='dodge',stat='identity')

```
