---
title: "Calculate drug enrichments"
author: "Sara gosline"
date: "2025-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(synapser)
library(ggplot2)
library(dplyr)
library(tidyr)

```

## Download drug stats
First we need to downlod the summary stats for each organoid 
```{r download, warning=FALSE}
synLogin()

fits <- readr::read_tsv(synGet('syn65471817')$path)

meta <- readxl::read_xlsx(syn$get('syn65595365')$path) |>
  tidyr::separate(Specimen,into = c('Patient','Tumor'),sep = '_',
                  remove = FALSE)

global <- readr::read_csv(synapser::synGet('syn65599827')$path)
phospho <- readr::read_csv(synapser::synGet('syn65598472')$path)
allcors <- readr::read_csv(synapser::synGet('syn66313804')$path)

pcols <- c(NF00017='steelblue',NF0021='orange3',NF0019='burlywood3',
           NF0022='goldenrod',NF0018='olivedrab',NF0020='darkred')

```

now that we have the correlations we can group proteins and phosphosites and do enrichment

## Global enrichment
We choose a few pathways and do enrichment on each drug based on unimolar sensitivyt and IC50.

```{r global enrichment ,warning=FALSE, message=FALSE,error=FALSE}
library(WebGestaltR)

alldrugs <- unique(allcor$drug)

alldrugs <- c('Selumetinib','Onalespib','Digoxin','Venetoclax')
gcors <- allcors |> subset(data == 'global')
go_global <- do.call(rbind,lapply(alldrugs,function(x){
  
  gprots <- subset(gcors,drug == x) |>
    dplyr::select(genes = 'gene',scores = 'pCor') |>
 as.data.frame()
  print(x)
  
  try(WebGestaltR(enrichMethod='GSEA',
                  minNum=3,
                  enrichDatabase='geneontology_Biological_Process',
                  fdrThr=1.0,
              interestGene=gprots,organism='hsapiens',
              interestGeneType='genesymbol',isOutput=FALSE)|>
    mutate(drug=x))

}))

subset(go_global,FDR<0.001)|>group_by(drug)|>summarize(n())


kegg_global<-do.call(rbind,lapply(alldrugs,function(x){
  
  gprots<-subset(gcors,drug==x)|>
    dplyr::select(genes='gene',scores='pCor')|>
 as.data.frame()
  print(x)
  
  try(WebGestaltR(enrichMethod='GSEA',enrichDatabase='pathway_KEGG',fdrThr=1.0,
              interestGene=gprots,minNum=3,
              organism='hsapiens',interestGeneType='genesymbol',
              isOutput=FALSE)|>
    mutate(drug=x))

}))

subset(kegg_global,FDR<0.001)|>group_by(drug)|>summarize(n())

```




Next we can visualize the results

```{r plot go/kegg, warning=FALSE}

library(ggplot2)
paths <- subset(kegg_global,FDR<0.001)

subset(kegg_global,description%in%paths$description) |>
  subset(FDR<0.001) |>
  ggplot(aes(x=description,y=drug,
             size=1-FDR,col=normalizedEnrichmentScore)) + 
  geom_point() + coord_flip()

paths<-subset(go_global,FDR<0.0001)

subset(go_global,description%in%paths$description)|>
  subset(FDR<0.00001)|>+s
  ggplot(aes(x=description,y=drug,size=1-FDR,col=normalizedEnrichmentScore))+geom_point()+coord_flip()+scale_color_manual(values=scale_col)
  
  ```



## Kinase enrichmnet
We can now look at kinase enrichment for each phosphosite

```{r kinase enrichment}

pcors<-subset(allcors,data=='phospho')
kin_enrich<-do.call(rbind,lapply(alldrugs,function(x){
  
  gprots<-subset(pcors,drug==x)|>
    dplyr::select(genes='gene',scores='pCor')|>
 as.data.frame()
  print(x)
  
  try(WebGestaltR(enrichMethod='GSEA',enrichDatabase='network_Kinase_phosphosite',fdrThr=1.0,
              interestGene=gprots,minNum=3,organism='hsapiens',interestGeneType='genesymbol',isOutput=FALSE)|>
    mutate(drug=x))

}))
```
