---
title: "Evaluate drug stats"
author: "Sara gosline"
date: "2025-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(synapser)
library(ggplot2)
library(dplyr)
library(tidyr)


```

## Pull processed files from previous markdowns
We have already run the previous scripts and stored data on Synapse


```{r pull files}

synLogin()
fits
glong
plong

```

## Format protein data to collect correlation values

Do simple correlations to identify putative trends in the data.

```{r make matrices of proteins}


shared<-intersect(fits$improve_sample_id,glong$Specimen)
print(paste('Found',length(shared),'shared samples'))

glob_dat<-glong|>ungroup()|>
  subset(Specimen%in%shared)|>
  dplyr::select(Specimen,Genes,meanAbundance)|>
  tidyr::pivot_wider(names_from='Genes',values_from='meanAbundance',values_fill=0,values_fn=mean)|>
  tibble::column_to_rownames('Specimen')
  

phos_dat<-plong|>ungroup()|>
  subset(Specimen%in%shared)|>
  dplyr::select(Specimen,site,meanAbundance)|>
  tidyr::pivot_wider(names_from='site',values_from='meanAbundance',values_fill=0,values_fn=mean)|>
  tibble::column_to_rownames('Specimen')


```

### Correlate uM viabilities with protein/phosphosite expression

Look at uM viabilities and see what we can find that correlates with expression

```{r find common samples, warning=FALSE,error=FALSE}


## a full join might be a challenge, maybe just take two matrices
drug_dat <- fits|>subset(dose_response_metric=='uM_viability')|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fn=mean)|>
  tibble::column_to_rownames('improve_sample_id')

gres<-cor(drug_dat[shared,],glob_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
  as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(glob_dat)),names_to='gene',values_to='pCor')|>
  arrange(desc(pCor))

##now lets try to get significance

gsig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(glob_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],glob_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(fdr=p.adjust(unlist(corp),method='fdr'))
  }))|>
  as.data.frame()|>
  mutate(drug=unlist(drug))|>
  mutate(gene=unlist(gene))

fullcors<-gres|>left_join(data.frame(gsig))|>mutate(data='global')

pres<-cor(drug_dat[shared,],phos_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
    as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(phos_dat)),names_to='site',values_to='pCor')|>
  arrange(desc(pCor))

##now lets look at correlations

psig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(phos_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],phos_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,
        silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(fdr=p.adjust(unlist(corp),method='fdr'))
  }))|>as.data.frame()

fullpcors<-pres|>rename(gene='site')|>left_join(data.frame(psig))|>mutate(data='phospho')

#combine all correlations
allcor<-rbind(fullcors,fullpcors)|>
  mutate(direction=ifelse(pCor<0,'neg','pos'))

readr::write_csv(allcor,'globalPhosphoDrugCors.csv')
synapser::synStore(synapser::File('globalPhosphoDrugCors.csv',parentId='syn66313784'))
##lets count the correlations and plot

corsummary<-allcor|>subset(fdr<0.05)|>
  group_by(drug,data,direction)|>
  summarize(features=n(),meanCor=mean(pCor))
corsummary|>
  subset(features>1)|>
  ggplot(aes(x=data,y=features,fill=drug))+facet_grid(~direction)+geom_bar(position='dodge',stat='identity')

ggsave('correlatedFeaturesFDR0.05.pdf')

```
Now that we have the correlations we can cherry pick some examples to showcase

```{r plot cors}

plotCors <- function(features,druglist,dataType='global'){
  ##subset a list of features and drugs and plot those in a graph
  require(ggplot2)
  if(dataType=='global'){
    ptab<-glong|>dplyr::rename(feature='Genes')
  }else{
    ptab<-plong|>dplyr::rename(feature='site')
  }
  dtab<-fits|>
    subset(dose_response_metric=='uM_viability')|>
    dplyr::rename(Specimen='improve_sample_id',Drug='improve_drug_id')|>
    subset(Drug%in%druglist)
  
  
  ftab<-features|>left_join(ptab)|>left_join(dtab)|>
    subset(!is.na(Drug))
  
  ftab|>ggplot(aes(x=meanAbundance,y=dose_response_value,col=Patient,size=1))+
    geom_point()+
    facet_grid(feature~Drug)+
    scale_color_manual(values=pcols)
}

druglist<-c('Onalespib')
features<-subset(allcor,drug%in%druglist)|>
  subset(fdr<0.1)|>
  subset(data=='phospho')

plotCors(rename(features,feature='gene'),druglist)
ggsave('onalespibFDR0.05Cors.pdf',height=5)

```

now we can visualize correlations
```{r check out HSP90s}

hsps<-unique(glong$Genes[grep('^HSP',glong$Genes)])

cor_hsps<-subset(allcor,gene%in%hsps)|>subset(drug=='Onalespib')|>subset(corp<0.05)
print(paste('measured',length(hsps),'HSPs in global data of which', nrow(cor_hsps),' are correlated with Onalespib'))

plotCors(rename(cor_hsps,feature='gene'),c('Onalespib'))

ggsave('hspCorsOna.pdf',width=8,height=14)

hspp<-unique(plong$site[grep('^HSP',plong$site)])
cor_hspps<-subset(allcor,gene%in%hspp)|>subset(drug=='Onalespib')|>subset(corp<0.01)
print(paste('measured',length(hspp),'HSPs in phospho data of which',nrow(cor_hspps),' are correlated with Onalespib'))


```

### Now lets look only at IC50 values

There are a few drugs for which we have IC50 values

```{r check ic50 cors,warning=FALSE,error=FALSE}

ifits<-subset(fits,dose_response_metric=='fit_ic50')

shared<-intersect(ifits$improve_sample_id,glong$Specimen)
print(paste('Found',length(shared),'shared samples'))

## a full join might be a challenge, maybe just take two matrices
drug_dat <- ifits|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fn=mean)|>
  tibble::column_to_rownames('improve_sample_id')

gres<-cor(drug_dat[shared,],glob_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
  as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(glob_dat)),names_to='gene',values_to='pCor')|>
  arrange(desc(pCor))

##now lets try to get significance

gsig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(glob_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],glob_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(fdr=p.adjust(unlist(corp),method='fdr'))
  }))|>
  as.data.frame()|>
  mutate(drug=unlist(drug))|>
  mutate(gene=unlist(gene))

fullcors<-gres|>left_join(data.frame(gsig))|>mutate(data='global')

pres<-cor(drug_dat[shared,],phos_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
    as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(phos_dat)),names_to='site',values_to='pCor')|>
  arrange(desc(pCor))

##now lets look at correlations

psig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(phos_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],phos_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(fdr=p.adjust(unlist(corp),method='fdr'))
  }))|>as.data.frame()

fullpcors<-pres|>rename(gene='site')|>left_join(data.frame(psig))|>mutate(data='phospho')

#combine all correlations
allcor<-rbind(fullcors,fullpcors)|>
  mutate(direction=ifelse(pCor<0,'neg','pos'))


##lets count the correlations and plot

corsummary<-allcor|>subset(fdr<0.1)|>
  group_by(drug,data,direction)|>
  summarize(features=n(),meanCor=mean(pCor))

corsummary|>
  #subset(features>1)|>
  ggplot(aes(x=data,y=features,fill=drug))+
  facet_grid(~direction)+
  geom_bar(position='dodge',stat='identity')

```

Again we have onalespib with numerous significantly correlated proteins, and one phosphosite for digoxin showing up .

```{r plot individual sites}

druglist<-c('Onalespib')
features<-subset(allcor,drug%in%druglist)|>
  subset(fdr<0.05)|>
  subset(data=='global')

plotCors(rename(features,feature='gene'),druglist)


druglist<-c('Digoxin')
features<-subset(allcor,drug%in%druglist)|>
  subset(fdr<0.1)|>
  subset(data=='phospho')

#plotCors(rename(features,feature='gene'),druglist,data='phospho')



```

Now we can check the HSP proteins directly

```{r HSP correlation}
hsps<-unique(glong$Genes[grep('^HSP',glong$Genes)])

cor_hsps<-subset(allcor,gene%in%hsps)|>subset(drug=='Onalespib')|>subset(corp<0.05)

print(paste('measured',length(hsps),'HSPs in global data of which', nrow(cor_hsps),' are correlated with Onalespib'))

plotCors(rename(cor_hsps,feature='gene'),'Onalespib')

hspp<-unique(plong$site[grep('^HSP',plong$site)])
cor_hspps<-subset(allcor,gene%in%hspp)|>subset(drug=='Onalespib')|>subset(corp<0.1)

print(paste('measured',length(hspp),'HSPs in phospho data of which',nrow(cor_hspps),' are correlated with Onalespib'))
```

The IC50 result is similar to the viability. Now we can check AUC

```{r auc hsp check, error=FALSE, warning=FALSE}

ifits<-subset(fits,dose_response_metric=='auc')

shared<-intersect(ifits$improve_sample_id,glong$Specimen)
print(paste('Found',length(shared),'shared samples'))

## a full join might be a challenge, maybe just take two matrices
drug_dat <- ifits|>
  dplyr::select(improve_sample_id,improve_drug_id,dose_response_value)|>
  tidyr::pivot_wider(names_from='improve_drug_id',values_from='dose_response_value',values_fn=mean)|>
  tibble::column_to_rownames('improve_sample_id')

gres<-cor(drug_dat[shared,],glob_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
  as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(glob_dat)),names_to='gene',values_to='pCor')|>
  arrange(desc(pCor))

##now lets try to get significance

gsig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(glob_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],glob_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>
  as.data.frame()|>
  mutate(drug=unlist(drug))|>
  mutate(gene=unlist(gene))

fullcors<-gres|>left_join(data.frame(gsig))|>mutate(data='global')

pres<-cor(drug_dat[shared,],phos_dat[shared,],use='pairwise.complete.obs',method='pearson')|>
    as.data.frame()|>
    tibble::rownames_to_column('drug')|>
  tidyr::pivot_longer(cols=2:(1+ncol(phos_dat)),names_to='site',values_to='pCor')|>
  arrange(desc(pCor))

##now lets look at correlations

psig<-do.call(rbind,lapply(colnames(drug_dat),function(x){
  do.call(rbind,lapply(colnames(phos_dat),function(y){
    pval<-1.0
    try(pval<-cor.test(drug_dat[shared,x],phos_dat[shared,y],use='pairwise.complete.obs',method='pearson')$p.value,silent=TRUE)
    return(c(corp=pval,drug=x,gene=y))
    }
    ))|>as.data.frame()|>
    mutate(bh_p=p.adjust(unlist(corp),method='BH'))
  }))|>as.data.frame()

fullpcors<-pres|>rename(gene='site')|>left_join(data.frame(psig))|>mutate(data='phospho')

#combine all correlations
allcor<-rbind(fullcors,fullpcors)|>
  mutate(direction=ifelse(pCor<0,'neg','pos'))


##lets count the correlations and plot

corsummary<-allcor|>subset(bh_p<0.1)|>
  group_by(drug,data,direction)|>
  summarize(features=n(),meanCor=mean(pCor))

corsummary|>
  #subset(features>1)|>
  ggplot(aes(x=data,y=features,fill=drug))+facet_grid(~direction)+geom_bar(position='dodge',stat='identity')

print(corsummary)

```
```{r HSP correlation again}
hsps<-unique(glong$Genes[grep('^HSP',glong$Genes)])

cor_hsps<-subset(allcor,gene%in%hsps)|>s
ubset(drug=='Onalespib')|>subset(corp<0.05)
print(paste('measured',length(hsps),'HSPs in global data of which', nrow(cor_hsps),' are correlated with Onalespib'))
cor_hsps

plotCors(rename(cor_hsps, feature='gene'),'Onalespib')

hspp<-unique(plong$site[grep('^HSP',plong$site)])
cor_hspps<-subset(allcor,gene%in%hspp)|>subset(drug=='Onalespib')|>subset(corp<0.1)
print(paste('measured',length(hspp),'HSPs in phospho data of which',nrow(cor_hspps),' are correlated with Onalespib'))
```
