---
title: "Targeted Biomarker Discovery for cNF Organoids"
author: "Jeremy Jacobson"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: flatly
---

# Overview

This notebook performs multi-strategy biomarker discovery for 7 drugs of interest in cNF organoids:
**Onalespib, Alvocidib, Mirdametinib, Selumetinib, Fimepinostat, Copanlisib, and Digoxin**.

## Approach

Standard FDR-corrected Spearman correlation finds very few significant hits at n=20 with 8,248 proteins (that pass the FDR filter of .25).
Instead, we use three complementary statistical strategies and require biomarkers to pass 2 or more:

1. **Spearman correlation** across all 20 samples
2. **Extreme responder analysis** (Wilcoxon rank-sum, top 5 vs bottom 5)
3. **Within-patient pairs** (paired Spearman, 6 patients with multiple tumors)

Proteins passing all three are designated **triple-hits [3s]** — the highest-confidence tier.
Proteins passing 2 strategies (shared across drugs) are designated **2-strategy recurrent [2s]**.

All analysis uses **global proteomics only** (no RNA or phospho).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 10,
  fig.height = 6,
  out.width  = "95%",
  cache      = TRUE
)

```

# Setup & Data Loading

```{r setup2, results='hide'}
source("../source/00_cNF_helper_code.R")
source("../source/02_analyze_modality_correlations.R")
source("../source/03_leapr_biomarker.R")

library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(readr)
library(knitr)

dir.create("biomarker_figs",   showWarnings = FALSE, recursive = TRUE)
dir.create("biomarker_tables", showWarnings = FALSE, recursive = TRUE)
```

```{r drugs_of_interest}
drugs_of_interest <- c("Onalespib", "Alvocidib", "Mirdametinib", "Selumetinib",
                       "Fimepinostat", "Copanlisib", "Digoxin")
```

```{r load_data, results='hide'}
# Drug viability
drugs <- readr::read_tsv(synGet("syn69947322")$path)

# Global proteomics (batch corrected)
glong <- readr::read_csv(synGet("syn70078416")$path)
```

```{r match_drug_names}
all_drug_ids  <- unique(drugs$improve_drug_id)
drug_id_upper <- toupper(all_drug_ids)
doi_upper     <- toupper(drugs_of_interest)

matched_ids <- sapply(doi_upper, function(d) {
  hit <- all_drug_ids[drug_id_upper == d]
  if (length(hit) == 0) hit <- all_drug_ids[grepl(d, drug_id_upper, fixed = TRUE)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
})
drugs_matched <- matched_ids[!is.na(matched_ids)]
```

```{r build_matrices}
# Build full drug matrix (all 238 drugs)
drug_mat_all <- make_drug_matrix(drugs, metric = "uM_viability")

# Run global proteomics correlation to get feature matrix
corr_global <- suppressWarnings(analyze_modality_correlations(
  df_long     = glong,
  sample_col  = "Specimen",
  feature_col = "Gene",
  value_col   = "correctedAbundance",
  drug_mat    = drug_mat_all[, drugs_matched, drop = FALSE],
  outdir      = "biomarker_figs",
  fdr_thresh  = 0.25
))

# Shared samples + feature matrix for downstream analysis
fm          <- as.matrix(corr_global$feat_mat)
shared_val  <- intersect(rownames(drug_mat_all), rownames(fm))
patient_val <- sub("_.*$", "", shared_val)
```

---

# 1. Dataset Summary

## 1.1 Sample and protein counts

```{r sample_summary}
cat(sprintf("Samples with proteomics + drug data: %d\n", length(shared_val)))
cat(sprintf("Unique patients: %d\n", length(unique(patient_val))))
cat(sprintf("Patients with 2+ tumors (paired analysis): %d\n",
            sum(table(patient_val) >= 2)))
cat(sprintf("Total proteins measured: %d\n", ncol(fm)))
cat(sprintf("Total drugs in screen: %d\n", ncol(drug_mat_all)))
cat(sprintf("Drugs tested in all 20 samples: %d\n",
            sum(colSums(!is.na(drug_mat_all[shared_val, ])) == length(shared_val))))

cat("\nSamples per patient:\n")
print(table(patient_val))
```

## 1.2 Drugs of interest: class, potency, variability

The 7 drugs of interest span 6 mechanism classes and the full range of variability seen in the
screen — from Copanlisib (SD = 0.044, lowest in the panel) to Mirdametinib (SD = 0.21, top decile).
The "N_Triple_Hits" column previews how many high-confidence biomarkers each drug yielded.

```{r doi_table}
drug_info <- data.frame(
  Drug = drugs_of_interest,
  Class = c("HSP90 inhibitor", "CDK inhibitor", "MEK inhibitor", "MEK inhibitor",
            "Dual PI3K/HDAC inhibitor", "PI3K inhibitor", "Cardiac glycoside"),
  Primary_Targets = c("HSP90a/b", "CDK9 (also CDK1/2/4/7)", "MEK1/2", "MEK1/2",
                      "PI3Ka/d + HDAC1/2/3/10", "PI3Ka/d", "Na+/K+-ATPase"),
  Mean_Viability = sapply(drugs_of_interest, function(d)
    round(mean(drug_mat_all[shared_val, d], na.rm = TRUE), 3)),
  SD_Viability = sapply(drugs_of_interest, function(d)
    round(sd(drug_mat_all[shared_val, d], na.rm = TRUE), 3)),
  N_Triple_Hits = c(13, 10, 21, 9, 19, 5, 6),
  stringsAsFactors = FALSE
) %>% arrange(SD_Viability)

knitr::kable(drug_info, caption = "Drugs of Interest: class, potency, and biomarker yield")
```

## 1.3 Sample-level drug response heatmap

Heatmap of viability across all 238 drugs × 20 samples grouped by patient. White cells = drug not
tested in that sample. Drugs are ordered by median viability (left = most potent). The visualization
shows two things at once: (1) a clear potent-drug cluster on the left side that kills nearly every
sample, and (2) substantial within-patient variation across the middle of the drug panel where
multiple-tumor patients (NF0017, NF0018, NF0019, NF0020, NF0025, NF0027) show different responses
between tumors — the foundation for the Strategy 3 paired analysis.

```{r drug_heatmap_all, fig.width=14, fig.height=14}
sample_order <- shared_val[order(patient_val, shared_val)]
patient_ordered <- sub("_.*$", "", sample_order)

heat_mat <- drug_mat_all[sample_order, ]
drug_medians <- apply(heat_mat, 2, median, na.rm = TRUE)
heat_mat <- heat_mat[, names(sort(drug_medians))]

patients_unique <- unique(patient_ordered)
gap_positions <- cumsum(table(patient_ordered)[patients_unique])
gap_positions <- gap_positions[-length(gap_positions)]

row_anno <- data.frame(Patient = patient_ordered)
rownames(row_anno) <- sample_order
ann_colors <- list(Patient = pcols[patients_unique])

pheatmap::pheatmap(
  heat_mat,
  color = colorRampPalette(c("steelblue", "lightyellow", "firebrick"))(100),
  na_col = "white",
  cluster_rows = FALSE, cluster_cols = FALSE,
  gaps_row = as.integer(gap_positions),
  annotation_row = row_anno, annotation_colors = ann_colors,
  show_colnames = FALSE,
  fontsize_row = 9, border_color = NA,
  main = sprintf("Drug Response: %d Drugs × %d Samples\nWhite = not tested. Drugs ordered by median viability (left = most potent).",
                 ncol(heat_mat), nrow(heat_mat)),
  legend_breaks = c(0.25, 0.5, 0.75, 1.0, 1.25),
  legend_labels = c("0.25", "0.5", "0.75", "1.0", "1.25\n\nViability"),
  filename = "biomarker_figs/final_drug_viability_heatmap_all.pdf",
  width = 14, height = 14
)
```

## 1.4 Drug landscape: potency vs variability

Each dot = 1 drug, x = mean viability, y = SD of viability. Drugs of interest are highlighted with
white-fill rings and bold labels; other drugs are colored by class. The plot demonstrates that
**drugs of interest span the full range of potency, variability, and mechanism** — Mirdametinib
sits at high SD/moderate potency, Copanlisib is at very low SD, Onalespib clusters with the rest
of the HSP90 inhibitors at moderate potency. This justifies treating each drug separately rather
than expecting a single biomarker to apply across all of them.

```{r drug_landscape, fig.width=12, fig.height=8}
drug_stats <- data.frame(
  drug = colnames(drug_mat_all),
  mean_viab = apply(drug_mat_all[shared_val, ], 2, mean, na.rm = TRUE),
  sd_viab   = apply(drug_mat_all[shared_val, ], 2, sd,   na.rm = TRUE),
  stringsAsFactors = FALSE
) %>% filter(is.finite(mean_viab), is.finite(sd_viab))

class_map <- list(
  "MEK" = c("Mirdametinib", "Selumetinib", "Trametinib", "Binimetinib",
            "Cobimetinib", "Avutometinib", "Nedometinib"),
  "HSP90" = c("Onalespib", "Tanespimycin", "Alvespimycin", "Ganetespib",
              "Luminespib", "Retaspimycin", "SNX-2112", "SNX-5422",
              "KW-2478", "CCT-018159", "Geldanamycin"),
  "CDK" = c("Alvocidib", "Dinaciclib", "Palbociclib", "Ribociclib",
            "Abemaciclib", "Trilaciclib", "Enitociclib", "Fadraciclib",
            "THZ1", "JNJ-7706621"),
  "PI3K/mTOR" = c("Copanlisib", "Alpelisib", "Omipalisib", "Apitolisib",
                  "Bimiralisib", "Eganelisib", "Umbralisib", "Duvelisib",
                  "NVP-BEZ235", "LY3023414", "Rapamycin", "Everolimus",
                  "Temsirolimus", "Sapanisertib", "Vistusertib"),
  "HDAC/BET" = c("Vorinostat", "Panobinostat", "Belinostat", "Romidepsin",
                 "Entinostat", "Fimepinostat", "Molibresib", "Pelabresib")
)

drug_stats$drug_class <- "Other"
for (cls in names(class_map)) {
  drug_stats$drug_class[drug_stats$drug %in% class_map[[cls]]] <- cls
}
drug_stats$drug_class[drug_stats$drug == "Digoxin"] <- "Cardiac glycoside"
drug_stats$is_doi <- drug_stats$drug %in% drugs_of_interest

class_levels <- c("MEK", "HSP90", "CDK", "PI3K/mTOR", "HDAC/BET", "Cardiac glycoside", "Other")
drug_stats$drug_class <- factor(drug_stats$drug_class, levels = class_levels)

class_colors <- c("MEK" = "#1C92AB", "HSP90" = "#C53030", "CDK" = "#DD6B20",
                  "PI3K/mTOR" = "#9F7AEA", "HDAC/BET" = "#276749",
                  "Cardiac glycoside" = "#1C5D8F", "Other" = "grey75")

drug_stats$label <- ifelse(drug_stats$is_doi, drug_stats$drug, "")

p_landscape <- ggplot(drug_stats, aes(x = mean_viab, y = sd_viab)) +
  geom_point(data = drug_stats %>% filter(drug_class == "Other"),
             color = "grey75", alpha = 0.5, size = 2) +
  geom_point(data = drug_stats %>% filter(drug_class != "Other" & !is_doi),
             aes(color = drug_class), size = 2.8, alpha = 0.85) +
  geom_point(data = drug_stats %>% filter(is_doi),
             aes(color = drug_class), size = 4.5, alpha = 0.95,
             stroke = 1.2, shape = 21, fill = "white") +
  geom_point(data = drug_stats %>% filter(is_doi),
             aes(color = drug_class), size = 3, alpha = 0.95) +
  geom_text_repel(data = drug_stats %>% filter(is_doi),
                  aes(label = label), size = 3.5, fontface = "bold",
                  max.overlaps = 20, segment.color = "grey40",
                  box.padding = 0.6, color = "grey20") +
  scale_color_manual(values = class_colors, name = "Drug Class",
                     breaks = c("MEK", "HSP90", "CDK", "PI3K/mTOR",
                                "HDAC/BET", "Cardiac glycoside")) +
  scale_x_continuous(limits = c(0, 1.25), breaks = seq(0, 1.25, 0.25)) +
  scale_y_continuous(limits = c(0, 0.25), breaks = seq(0, 0.25, 0.05)) +
  geom_vline(xintercept = 1.0, linetype = "dashed", color = "grey60", linewidth = 0.3) +
  labs(
    title = "Drug Landscape: Potency vs Variability",
    subtitle = "Drugs of interest span the full range of potency, variability, and mechanism",
    x = "Mean Viability (lower = more potent)",
    y = "SD of Viability (response variability)"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right",
        legend.title = element_text(face = "bold"),
        panel.grid.minor = element_blank())

print(p_landscape)
ggsave("biomarker_figs/final_drug_landscape_scatter.pdf", p_landscape, width = 12, height = 8)
```

## 1.5 Triple-hit summary per drug

Compiled list of all 83 triple-hit protein-drug pairs across the 7 drugs of interest. Each row is
a protein that passed all three statistical strategies for the given drug, with rho and direction
(resistance = positive rho, sensitivity = negative rho).

```{r triple_hit_table}
combined <- do.call(rbind, lapply(drugs_of_interest, function(d) {
  hits <- triple_detail %>% filter(drug == d)
  if (nrow(hits) == 0) return(NULL)
  hits$rho <- sapply(hits$feature, function(p) {
    if (p %in% colnames(fm)) {
      round(suppressWarnings(cor(drug_mat_all[shared_val, d], fm[shared_val, p],
                                  method = "spearman", use = "pairwise.complete.obs")), 3)
    } else { NA_real_ }
  })
  data.frame(
    Drug = d, Protein = hits$feature, Rho = hits$rho,
    Direction = ifelse(hits$rho > 0, "Resistance", "Sensitivity"),
    stringsAsFactors = FALSE
  )
})) %>% arrange(Drug, desc(abs(Rho)))

cat(sprintf("All triple-hit [3s] biomarkers: %d protein-drug pairs across %d drugs\n\n",
            nrow(combined), length(unique(combined$Drug))))
knitr::kable(combined, caption = "All triple-hit biomarkers per drug, sorted by |rho|")
```

---

# 2. Multi-Strategy Discovery Approach

## 2.1 Why three strategies are necessary

Standard FDR-corrected Spearman with n=20 across 8,248 proteins requires |rho| ≥ 0.75 to survive
correction. Very few single protein-drug pair achieves this bar. The multi-strategy consensus approach finds
biomarkers that any single test would miss by requiring concordance across three independent tests
that ask different questions of the same data: continuous correlation across all samples, extreme
responder differences, and within-patient consistency.

```{r strategy_helpers}
# Helper: extreme responder analysis for one drug
run_extreme_responders <- function(drug_id, drug_mat, feat_mat, n_extreme = 5) {
  shared <- intersect(rownames(drug_mat), rownames(feat_mat))
  if (!(drug_id %in% colnames(drug_mat))) return(NULL)
  viab <- drug_mat[shared, drug_id]; viab <- viab[is.finite(viab)]
  viab_sorted <- sort(viab)
  sens <- names(viab_sorted)[1:n_extreme]
  res  <- names(viab_sorted)[(length(viab_sorted) - n_extreme + 1):length(viab_sorted)]

  results <- do.call(rbind, lapply(colnames(feat_mat), function(f) {
    s_vals <- feat_mat[sens, f]; r_vals <- feat_mat[res, f]
    if (all(is.na(s_vals)) || all(is.na(r_vals))) return(NULL)
    if (sd(c(s_vals, r_vals), na.rm = TRUE) == 0) return(NULL)
    wt <- suppressWarnings(wilcox.test(r_vals, s_vals, exact = FALSE))
    data.frame(drug = drug_id, feature = f, pval = wt$p.value,
               mean_sens = mean(s_vals, na.rm = TRUE),
               mean_res = mean(r_vals, na.rm = TRUE),
               direction = ifelse(mean(r_vals, na.rm = TRUE) > mean(s_vals, na.rm = TRUE),
                                  "up_in_resistant", "up_in_sensitive"),
               stringsAsFactors = FALSE)
  }))
  if (is.null(results)) return(NULL)
  results$fdr <- p.adjust(results$pval, method = "BH")
  results %>% arrange(pval)
}

# Helper: within-patient paired Spearman on deltas
run_within_patient <- function(drug_id, drug_mat, feat_mat, sample_info, multi_patients) {
  if (!(drug_id %in% colnames(drug_mat))) return(NULL)
  pairs <- do.call(rbind, lapply(multi_patients, function(pat) {
    samps <- sample_info$sample[sample_info$Patient == pat]
    viabs <- drug_mat[samps, drug_id]; viabs <- viabs[is.finite(viabs)]
    if (length(viabs) < 2) return(NULL)
    most_sens <- names(which.min(viabs)); most_res <- names(which.max(viabs))
    if (most_sens == most_res) return(NULL)
    data.frame(patient = pat, sensitive = most_sens, resistant = most_res,
               viab_diff = viabs[most_res] - viabs[most_sens],
               stringsAsFactors = FALSE)
  }))
  if (is.null(pairs) || nrow(pairs) < 3) return(NULL)

  feat_diffs <- sapply(colnames(feat_mat), function(f) {
    feat_mat[pairs$resistant, f] - feat_mat[pairs$sensitive, f]
  })
  results <- apply(feat_diffs, 2, function(fd) {
    if (all(is.na(fd)) || sd(fd, na.rm = TRUE) == 0) return(c(rho = NA, pval = NA))
    ct <- suppressWarnings(cor.test(pairs$viab_diff, fd, method = "spearman", exact = FALSE))
    c(rho = unname(ct$estimate), pval = unname(ct$p.value))
  })
  res_df <- as.data.frame(t(results)) %>% tibble::rownames_to_column("feature") %>%
    dplyr::mutate(fdr = p.adjust(pval, method = "BH")) %>%
    dplyr::arrange(pval)
  res_df$drug <- drug_id
  res_df
}

# Helper: format p-values for plot annotations
fmt_p <- function(p) {
  if (length(p) == 0 || !is.finite(p)) return("NA")
  if (p < 0.001) return("<0.001")
  sprintf("%.3f", p)
}
```

## 2.2 Run all three strategies

```{r run_strategies, cache=TRUE}
# Strategy 1: Spearman correlation (already computed in corr_global)

# Strategy 2: Extreme responders
extreme_results <- list()
for (d in drugs_matched) {
  res <- run_extreme_responders(d, as.matrix(drug_mat_all), fm, n_extreme = 5)
  if (!is.null(res)) extreme_results[[d]] <- res
}
extreme_all <- bind_rows(extreme_results)

# Strategy 3: Within-patient pairs
sample_info <- data.frame(sample = shared_val, Patient = patient_val,
                          stringsAsFactors = FALSE)
multi_patients <- sample_info %>% dplyr::count(Patient) %>%
  filter(n >= 2) %>% pull(Patient)

within_patient_results <- list()
for (d in drugs_matched) {
  res <- run_within_patient(d, as.matrix(drug_mat_all), fm, sample_info, multi_patients)
  if (!is.null(res)) within_patient_results[[d]] <- res
}
```

## 2.3 Identify triple-hit and 2-strategy recurrent proteins

```{r identify_triples}
hits_correlation <- corr_global$cor_tbl %>%
  filter(drug %in% drugs_matched, pval < 0.05) %>%
  select(drug, feature) %>% mutate(strategy = "correlation")

hits_extreme <- extreme_all %>%
  filter(pval < 0.05) %>%
  select(drug, feature) %>% mutate(strategy = "extreme_responders")

hits_within <- bind_rows(within_patient_results) %>%
  filter(pval < 0.05) %>%
  select(drug, feature) %>% mutate(strategy = "within_patient")

all_hits <- bind_rows(hits_correlation, hits_extreme, hits_within)

recurrent <- all_hits %>%
  group_by(drug, feature) %>%
  summarise(n_strategies = n_distinct(strategy),
            strategies   = paste(unique(strategy), collapse = " + "),
            .groups = "drop") %>%
  arrange(desc(n_strategies), drug, feature)

triple_detail <- recurrent %>% filter(n_strategies == 3)

cat(sprintf("Triple-hit [3s] biomarkers: %d\n", nrow(triple_detail)))
cat(sprintf("2-strategy [2s] recurrent: %d\n", sum(recurrent$n_strategies == 2)))

write_csv(recurrent, "biomarker_tables/recurrent_hits_multi_strategy.csv")
write_csv(triple_detail, "biomarker_tables/triple_hit_biomarkers.csv")
```

---

# 3. Finding 1: HSP90 Chaperone Network (Onalespib + Cross-Drug)

## 3.1 Rationale

Onalespib yielded ST13 as a triple-hit biomarker. ST13 is a HSP70 co-chaperone that stabilizes
the HSP70-HSP90 machinery. Independent discoveries from Alvocidib (NUDCD3) and Digoxin (DNAJC8)
also identified co-chaperone proteins, suggesting a cross-drug chaperone resistance axis.
We validated the full 7-protein chaperone panel across all 11 HSP90 inhibitors in the screen.

## 3.2 ST13 vs Onalespib viability — all three strategies

ST13 abundance correlates strongly with Onalespib viability (rho = 0.77), and the 
three strategy p-values are annotated to make the multi-strategy concordance 
explicit: nominal Spearman p, unpaired Wilcoxon p (extreme responders), and paired
Spearman p (within-patient deltas). All three are significant at p < 0.05, qualifying ST13 as a
triple-hit. Higher ST13 → higher viability → less drug effect.

```{r st13_headline, fig.width=10, fig.height=7}
# Pull the three strategy p-values from the discovery objects
st13_cor_p <- corr_global$cor_tbl %>%
  filter(drug == "Onalespib", feature == "ST13") %>%
  pull(pval)
st13_extreme_p <- extreme_all %>%
  filter(drug == "Onalespib", feature == "ST13") %>%
  pull(pval)
st13_within_p <- within_patient_results[["Onalespib"]] %>%
  filter(feature == "ST13") %>%
  pull(pval)

df_st13_main <- data.frame(
  sample    = shared_val,
  Patient   = patient_val,
  viability = drug_mat_all[shared_val, "Onalespib"],
  ST13      = fm[shared_val, "ST13"]
) %>% filter(is.finite(viability), is.finite(ST13))

rho_st13_main <- cor(df_st13_main$ST13, df_st13_main$viability,
                     method = "spearman", use = "pairwise.complete.obs")

st13_label <- sprintf(
  "rho = %.2f\ncor p = %s\nWilcox p = %s\npaired p = %s",
  rho_st13_main,
  fmt_p(st13_cor_p),
  fmt_p(st13_extreme_p),
  fmt_p(st13_within_p)
)

p_st13_main <- ggplot(df_st13_main, aes(x = ST13, y = viability)) +
  geom_point(aes(color = Patient), size = 4, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = pcols) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
           label = st13_label, size = 3.8, color = "grey20", lineheight = 1.1) +
  labs(
    title = "ST13 (HSP70 Co-Chaperone / HSP70-HSP90 Adaptor) Predicts Onalespib Resistance",
    subtitle = "Triple-hit [3s] discovery: passes all three statistical strategies",
    x = "ST13 Protein Abundance (normalized)",
    y = "Onalespib Viability"
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")
print(p_st13_main)
ggsave("biomarker_figs/final_ST13_onalespib_headline.pdf", p_st13_main, width = 10, height = 7)
```

## 3.3 ST13 across 6 HSP90 inhibitors with all three strategy p-values

ST13 validates as a resistance marker across 5 structurally distinct HSP90 inhibitors beyond the
discovery drug Onalespib. Each panel shows the rho plus all three strategy p-values, demonstrating
that the validation isn't a single-statistic artifact - multiple HSP90 inhibitors show concordant
multi-strategy support for ST13 as a resistance biomarker.

```{r st13_multi_strategy, fig.width=16, fig.height=10}
top_hsp90 <- intersect(
  c("Onalespib", "SNX-2112", "SNX-5422", "Ganetespib", "Retaspimycin", "Tanespimycin"),
  colnames(drug_mat_all))

# Helper to compute all 3 strategies for one drug × ST13
compute_st13_stats <- function(drug_id, n_extreme = 5) {
  viab <- drug_mat_all[shared_val, drug_id]
  prot <- fm[shared_val, "ST13"]
  keep <- is.finite(viab) & is.finite(prot)
  v <- viab[keep]; p <- prot[keep]
  samp <- shared_val[keep]
  n <- length(v)

  s1 <- suppressWarnings(cor.test(p, v, method = "spearman", exact = FALSE))
  rho_val <- unname(s1$estimate); cor_p <- s1$p.value

  wilcox_p <- NA_real_
  if (n >= 2 * n_extreme) {
    ord <- order(v)
    sens <- ord[1:n_extreme]; res <- ord[(n - n_extreme + 1):n]
    wt <- suppressWarnings(wilcox.test(p[res], p[sens], exact = FALSE))
    wilcox_p <- wt$p.value
  }

  paired_p <- NA_real_
  patient <- sub("_.*$", "", samp)
  multi_pat <- unique(patient[duplicated(patient)])
  if (length(multi_pat) >= 3) {
    deltas <- do.call(rbind, lapply(multi_pat, function(pat) {
      idx <- which(patient == pat)
      if (length(idx) < 2) return(NULL)
      sens_i <- idx[which.min(v[idx])]; res_i <- idx[which.max(v[idx])]
      if (sens_i == res_i) return(NULL)
      data.frame(viab_diff = v[res_i] - v[sens_i],
                 prot_diff = p[res_i] - p[sens_i])
    }))
    if (!is.null(deltas) && nrow(deltas) >= 3) {
      pt <- suppressWarnings(cor.test(deltas$viab_diff, deltas$prot_diff,
                                      method = "spearman", exact = FALSE))
      paired_p <- pt$p.value
    }
  }
  list(rho = rho_val, cor_p = cor_p, wilcox_p = wilcox_p, paired_p = paired_p)
}

plots_st13 <- lapply(top_hsp90, function(d) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, d],
    ST13 = fm[shared_val, "ST13"]
  ) %>% filter(is.finite(viability), is.finite(ST13))

  stats <- compute_st13_stats(d)
  is_disc <- (d == "Onalespib")
  label_text <- sprintf(
    "rho = %.2f\ncor p = %s\nWilcox p = %s\npaired p = %s",
    stats$rho, fmt_p(stats$cor_p), fmt_p(stats$wilcox_p), fmt_p(stats$paired_p)
  )

  ggplot(df, aes(x = ST13, y = viability)) +
    geom_point(aes(color = Patient), size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             label = label_text, size = 3.2, color = "grey20", lineheight = 1.1) +
    labs(title = if (is_disc) paste0(d, " (discovery)") else d,
         x = "ST13 Abundance", y = "Viability") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(face = if (is_disc) "bold" else "plain"))
})

p_st13 <- wrap_plots(plots_st13, ncol = 3) +
  plot_annotation(
    title = "ST13 [3s] (HSP70 Co-Chaperone) Predicts Resistance Across 6 HSP90 Inhibitors",
    subtitle = "Triple-hit discovery in Onalespib, validated in 5 structurally distinct compounds. Each panel shows all 3 strategy p-values."
  )
print(p_st13)
ggsave("biomarker_figs/final_ST13_multi_hsp90.pdf", p_st13, width = 16, height = 10)
```

## 3.4 Chaperone network heatmap with discovery source annotation

The full 7-protein chaperone panel × 13 drugs (11 HSP90 inhibitors + Alvocidib + Digoxin). The
column color annotation shows where each protein was originally discovered: ST13 from Onalespib
(red), NUDCD3 from Alvocidib (orange), DNAJC8 from Digoxin (blue), and 4 multi-drug recurrent
proteins (grey). This visualization highlights that the chaperone network was independently
identified across three structurally unrelated drugs and validates most strongly within the HSP90
inhibitor class — proteostasis as a master regulator of drug response.

```{r chaperone_heatmap, fig.width=12, fig.height=9}
chap_genes <- c("HSP90AA1", "HSPA4", "HSPA8", "ST13", "NUDCD3", "CHORDC1", "DNAJC8")
chap_in <- intersect(chap_genes, colnames(fm))

chap_evidence <- c("HSP90AA1" = " [2s]", "HSPA4" = " [2s]", "HSPA8" = " [2s]",
                   "ST13" = " [3s]", "NUDCD3" = " [3s]",
                   "CHORDC1" = " [2s]", "DNAJC8" = " [3s]")

chap_discovery <- c("HSP90AA1" = "Multi-drug", "HSPA4" = "Multi-drug",
                    "HSPA8" = "Multi-drug", "ST13" = "Onalespib",
                    "NUDCD3" = "Alvocidib", "CHORDC1" = "Multi-drug",
                    "DNAJC8" = "Digoxin")

hsp90_drugs <- c("Onalespib", "Tanespimycin", "Alvespimycin", "Ganetespib",
                 "Luminespib", "Retaspimycin", "SNX-2112", "SNX-5422",
                 "KW-2478", "CCT-018159", "Geldanamycin")
extra_drugs <- c("Alvocidib", "Digoxin")
all_drugs_heatmap <- intersect(c(hsp90_drugs, extra_drugs), colnames(drug_mat_all))

cor_mat <- matrix(NA, nrow = length(all_drugs_heatmap), ncol = length(chap_in),
                  dimnames = list(all_drugs_heatmap, chap_in))
for (d in all_drugs_heatmap) {
  for (p in chap_in) {
    cor_mat[d, p] <- suppressWarnings(
      cor(drug_mat_all[shared_val, d], fm[shared_val, p],
          method = "spearman", use = "pairwise.complete.obs"))
  }
}

col_labels <- sapply(colnames(cor_mat), function(p) paste0(p, chap_evidence[p]))
colnames(cor_mat) <- col_labels

drug_class_anno <- data.frame(
  Class = ifelse(rownames(cor_mat) %in% hsp90_drugs, "HSP90 inhibitor",
          ifelse(rownames(cor_mat) == "Alvocidib", "CDK9 inhibitor",
          ifelse(rownames(cor_mat) == "Digoxin", "Cardiac glycoside", "Other")))
)
rownames(drug_class_anno) <- rownames(cor_mat)

discovery_source <- data.frame(
  Discovery = sapply(chap_in, function(p) chap_discovery[p])
)
rownames(discovery_source) <- col_labels

ann_colors_chap <- list(
  Class = c("HSP90 inhibitor" = "firebrick", "CDK9 inhibitor" = "darkorange",
            "Cardiac glycoside" = "steelblue"),
  Discovery = c("Onalespib" = "firebrick", "Alvocidib" = "darkorange",
                "Digoxin" = "steelblue", "Multi-drug" = "grey50")
)

pheatmap::pheatmap(
  cor_mat,
  color = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
  breaks = seq(-1, 1, length.out = 101),
  display_numbers = TRUE, number_format = "%.2f", fontsize_number = 8,
  annotation_row = drug_class_anno,
  annotation_col = discovery_source,
  annotation_colors = ann_colors_chap,
  cluster_rows = TRUE, cluster_cols = TRUE,
  legend_breaks = c(-1, -0.5, 0, 0.5, 1),
  legend_labels = c("-1", "-0.5", "0", "0.5", "1\n\nSpearman\nrho"),
  main = "Chaperone Network: Discovered Across Drug Classes, Validated for HSP90 Inhibitors\n[3s] = triple-hit    [2s] = 2-strategy recurrent    Column color = discovery drug",
  fontsize = 11, angle_col = 45,
  filename = "biomarker_figs/final_hsp90_chaperone_heatmap_annotated.pdf",
  width = 9, height = 9
)
```

## 3.5 Pan-drug chaperone score waterfall

Three-panel waterfall ranking all 238 drugs by their correlation with the chaperone score
(mean of HSP90AA1 + ST13 + HSPA8). Top: Spearman rho. Middle: mean viability. Bottom: SD of
viability. HSP90 inhibitors (red) dominate the top of the rho ranking — chaperone capacity is
the primary HSP90 inhibitor resistance axis. The middle and bottom panels add context by
identifying anti-chaperone potent drugs (combination candidates) and highlighting variability.

```{r chaperone_waterfall, fig.width=16, fig.height=12}
chap_score_genes <- c("HSP90AA1", "ST13", "HSPA8")
chap_score_in <- intersect(chap_score_genes, colnames(fm))
chap_score <- rowMeans(fm[shared_val, chap_score_in, drop = FALSE], na.rm = TRUE)

chap_drug_df <- data.frame(
  drug = colnames(drug_mat_all),
  rho = sapply(colnames(drug_mat_all), function(d) {
    suppressWarnings(cor(drug_mat_all[shared_val, d], chap_score,
                         method = "spearman", use = "pairwise.complete.obs"))
  }),
  mean_viab = sapply(colnames(drug_mat_all), function(d)
    mean(drug_mat_all[shared_val, d], na.rm = TRUE)),
  sd_viab = sapply(colnames(drug_mat_all), function(d)
    sd(drug_mat_all[shared_val, d], na.rm = TRUE)),
  stringsAsFactors = FALSE
) %>% filter(is.finite(rho)) %>%
  arrange(desc(rho)) %>%
  mutate(rank = row_number())

chap_drug_df$is_hsp90 <- chap_drug_df$drug %in% hsp90_drugs

chap_drug_df$label_top <- ifelse(chap_drug_df$rank <= 15 | chap_drug_df$is_hsp90,
                                  chap_drug_df$drug, "")

lowest_neg <- chap_drug_df %>% filter(rho < 0) %>%
  arrange(mean_viab) %>% slice_head(n = 3) %>% pull(drug)
lowest_all <- chap_drug_df %>%
  arrange(mean_viab) %>% slice_head(n = 3) %>% pull(drug)
label_mid_drugs <- unique(c(lowest_neg, lowest_all))
chap_drug_df$label_mid <- ifelse(chap_drug_df$drug %in% label_mid_drugs | chap_drug_df$is_hsp90,
                                  chap_drug_df$drug, "")

top_var <- chap_drug_df %>% arrange(desc(sd_viab)) %>% slice_head(n = 5) %>% pull(drug)
top_var_neg <- chap_drug_df %>% filter(rho < 0) %>%
  arrange(desc(sd_viab)) %>% slice_head(n = 3) %>% pull(drug)
least_var <- chap_drug_df %>% arrange(sd_viab) %>% slice_head(n = 3) %>% pull(drug)
label_bot_drugs <- unique(c(top_var, top_var_neg, least_var))
chap_drug_df$label_bot <- ifelse(chap_drug_df$drug %in% label_bot_drugs | chap_drug_df$is_hsp90,
                                  chap_drug_df$drug, "")

chap_drug_df$drug_f <- factor(chap_drug_df$drug, levels = chap_drug_df$drug)

shared_theme <- theme_bw(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        plot.margin = margin(2, 10, 0, 10))

p_top <- ggplot(chap_drug_df, aes(x = drug_f, y = rho, fill = is_hsp90)) +
  geom_col(width = 1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.3) +
  geom_text(data = chap_drug_df %>% filter(label_top != ""),
            aes(label = label_top), angle = 90, hjust = -0.05, size = 2.2,
            show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "grey75", "TRUE" = "firebrick"),
                    labels = c("Other drugs", "HSP90 inhibitors")) +
  labs(title = "Chaperone Score (mean of HSP90AA1, ST13, HSPA8) vs All Drugs",
       y = "Spearman rho\n(chaperone association)", fill = NULL) +
  shared_theme +
  theme(legend.position = "top") +
  coord_cartesian(ylim = c(-0.5, 1.0), clip = "off")

p_mid <- ggplot(chap_drug_df, aes(x = drug_f, y = mean_viab, fill = is_hsp90)) +
  geom_col(width = 1) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  geom_text(data = chap_drug_df %>% filter(label_mid != ""),
            aes(label = label_mid), angle = 90, hjust = -0.05, size = 2.2,
            show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "grey75", "TRUE" = "firebrick")) +
  labs(y = "Mean Viability\n(lower = more potent)") +
  shared_theme +
  theme(legend.position = "none") +
  coord_cartesian(ylim = c(0, 1.5), clip = "off")

p_bot <- ggplot(chap_drug_df, aes(x = drug_f, y = sd_viab, fill = is_hsp90)) +
  geom_col(width = 1) +
  geom_text(data = chap_drug_df %>% filter(label_bot != ""),
            aes(label = label_bot), angle = 90, hjust = -0.05, size = 2.2,
            show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "grey75", "TRUE" = "firebrick")) +
  labs(x = "Drugs (ranked by chaperone rho: left = most resistant, right = most sensitive)",
       y = "SD of Viability\n(response variability)") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 10, 5, 10)) +
  coord_cartesian(ylim = c(0, 0.35), clip = "off")

p_triple <- p_top / p_mid / p_bot + plot_layout(heights = c(1.2, 0.8, 0.6))
print(p_triple)
ggsave("biomarker_figs/final_chaperone_triple_waterfall.pdf", p_triple, width = 16, height = 12)
```

---

# 4. Finding 2: MEK Inhibitor Sensitivity Biomarkers

## 4.1 Rationale

Mirdametinib and Selumetinib share 9 proteins at the 2-strategy level that split cleanly into
4 resistance markers and 5 sensitivity markers. The sensitivity markers — FECH, AS3MT, STIM1,
PPP1R1B, ABHD11 — validate class-wide across all 6 MEK inhibitors tested, while the resistance
markers are more drug-specific. FECH (ferrochelatase, terminal heme biosynthesis) is the cleanest
finding and connects to a convergent mitochondrial theme.

## 4.2 MEK class-wide biomarker heatmap (resistance vs sensitivity)

Two-panel heatmap showing the 4 resistance markers (left) and 5 sensitivity markers (right) across
6 MEK inhibitors. The sensitivity panel shows consistent negative rho values across the entire
class — these biomarkers generalize. The resistance panel is messier with mixed colors, indicating
those proteins are more Mirdametinib/Selumetinib-specific. The takeaway: **MEK sensitivity
biomarkers generalize across the class; resistance markers do not**.

```{r mek_split_heatmap, fig.width=14, fig.height=6}
mek_resist <- c("ABCF1", "TUBA1C", "PPP2R1A", "RETN")
mek_sensit <- c("FECH", "AS3MT", "STIM1", "PPP1R1B", "ABHD11")

mek_resist_evidence <- rep(" [2s]", length(mek_resist)); names(mek_resist_evidence) <- mek_resist
mek_sensit_evidence <- rep(" [2s]", length(mek_sensit)); names(mek_sensit_evidence) <- mek_sensit

mek_drugs_6 <- intersect(c("Mirdametinib", "Selumetinib", "Trametinib",
                            "Binimetinib", "Cobimetinib", "Avutometinib"),
                         colnames(drug_mat_all))

mek_drug_labels <- c(
  "Mirdametinib" = "Mirdametinib (MEK1/2)",
  "Selumetinib"  = "Selumetinib (MEK1/2)",
  "Trametinib"   = "Trametinib (MEK1/2)",
  "Binimetinib"  = "Binimetinib (MEK1/2)",
  "Cobimetinib"  = "Cobimetinib (MEK1)",
  "Avutometinib" = "Avutometinib (RAF/MEK)"
)

build_mek_cor <- function(proteins, evidence_map, label) {
  do.call(rbind, lapply(mek_drugs_6, function(d) {
    do.call(rbind, lapply(proteins, function(p) {
      if (!(p %in% colnames(fm))) return(NULL)
      rho <- suppressWarnings(cor(drug_mat_all[shared_val, d], fm[shared_val, p],
                                   method = "spearman", use = "pairwise.complete.obs"))
      p_marked <- paste0(p, evidence_map[p])
      data.frame(drug = mek_drug_labels[d], protein = p_marked, rho = rho,
                 group = label, stringsAsFactors = FALSE)
    }))
  }))
}

mek_all_cors <- rbind(
  build_mek_cor(intersect(mek_resist, colnames(fm)), mek_resist_evidence,
                "Resistance\n(higher \u2192 less effect)"),
  build_mek_cor(intersect(mek_sensit, colnames(fm)), mek_sensit_evidence,
                "Sensitivity\n(higher \u2192 more effect)")
)

drug_order_mek <- c("Mirdametinib (MEK1/2)", "Selumetinib (MEK1/2)", "Trametinib (MEK1/2)",
                    "Binimetinib (MEK1/2)", "Cobimetinib (MEK1)", "Avutometinib (RAF/MEK)")
mek_all_cors$drug <- factor(mek_all_cors$drug, levels = rev(drug_order_mek))

p_mek <- ggplot(mek_all_cors, aes(x = protein, y = drug, fill = rho)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 3.3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-0.7, 0.7)) +
  facet_wrap(~ group, scales = "free_x") +
  labs(title = "MEK Inhibitor Class-Wide Biomarkers \u2014 Resistance vs Sensitivity",
       subtitle = "MEK sensitivity biomarkers generalize across the class; resistance markers do not",
       x = NULL, y = NULL, fill = "Spearman\nrho") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank(),
        strip.text = element_text(face = "bold", size = 11))
print(p_mek)
ggsave("biomarker_figs/final_mek_resist_sensit.pdf", p_mek, width = 14, height = 6)
```

## 4.3 FECH sensitivity scatter across all 6 MEK inhibitors

Per-drug scatter showing FECH abundance vs viability for each MEK inhibitor. Negative slope
(sensitivity) is consistent across all 6 drugs, confirming the heatmap finding. FECH is the
terminal enzyme of heme biosynthesis; higher FECH → more heme → more functional cytochromes →
more active mitochondrial electron transport → more sensitive to MEK blockade. This connects
directly to the Selumetinib mitochondrial signature in section 4.4.

```{r fech_scatter, fig.width=14, fig.height=8}
mek_class_labels <- c(
  "Mirdametinib" = "Mirdametinib\n(MEK1/2)",
  "Selumetinib"  = "Selumetinib\n(MEK1/2)",
  "Trametinib"   = "Trametinib\n(MEK1/2)",
  "Binimetinib"  = "Binimetinib\n(MEK1/2)",
  "Cobimetinib"  = "Cobimetinib\n(MEK1)",
  "Avutometinib" = "Avutometinib\n(RAF/MEK)"
)

df_fech <- do.call(rbind, lapply(mek_drugs_6, function(d) {
  data.frame(sample = shared_val, Patient = patient_val,
             drug = mek_class_labels[d],
             viability = drug_mat_all[shared_val, d],
             FECH = fm[shared_val, "FECH"], stringsAsFactors = FALSE)
})) %>% filter(is.finite(viability), is.finite(FECH))

ann_fech <- df_fech %>%
  group_by(drug) %>%
  summarise(rho = suppressWarnings(cor(FECH, viability, method = "spearman")),
            .groups = "drop") %>%
  mutate(label = sprintf("rho = %.2f", rho))

p_fech <- ggplot(df_fech, aes(x = FECH, y = viability)) +
  geom_point(aes(color = Patient), size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.6) +
  geom_text(data = ann_fech, aes(label = label), x = -Inf, y = Inf,
            hjust = -0.05, vjust = 1.5, size = 3.5, color = "grey20", inherit.aes = FALSE) +
  facet_wrap(~ drug, ncol = 3) +
  scale_color_manual(values = pcols) +
  labs(title = "Ferrochelatase (FECH) Predicts MEK Inhibitor Sensitivity Across the Class",
       subtitle = "Higher FECH \u2192 more heme \u2192 more active mitochondrial ETC \u2192 more sensitive",
       x = "FECH Abundance (normalized)", y = "Viability") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")
print(p_fech)
ggsave("biomarker_figs/final_FECH_sensitivity.pdf", p_fech, width = 14, height = 8)
```

## 4.4 Selumetinib mitochondrial translation signature

UQCRC1 (Complex III subunit), PARS2 (mitochondrial prolyl-tRNA synthetase), and MRPL2
(mitochondrial ribosomal protein L2) are all triple-hits for Selumetinib. Their averaged score
correlates strongly (rho = -0.59) with Selumetinib viability, converging with the FECH finding:
mitochondrially active tumors are more sensitive to MEK blockade. This is a coherent biological
signal across two independent biomarker discoveries (FECH from MEK class, mito translation from
Selumetinib triple-hits).

```{r selumetinib_mito, fig.width=9, fig.height=6}
mito_in <- intersect(c("UQCRC1", "PARS2", "MRPL2"), colnames(fm))
mito_score <- rowMeans(fm[shared_val, mito_in, drop = FALSE], na.rm = TRUE)

df_mito <- data.frame(
  sample = shared_val, Patient = patient_val,
  viability = drug_mat_all[shared_val, "Selumetinib"],
  mito_score = mito_score
) %>% filter(is.finite(viability), is.finite(mito_score))

rho_mito <- suppressWarnings(cor(df_mito$mito_score, df_mito$viability, method = "spearman"))

p_mito <- ggplot(df_mito, aes(x = mito_score, y = viability)) +
  geom_point(aes(color = Patient), size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = pcols) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
           label = sprintf("rho = %.2f (sensitivity)\nHigher mitochondrial translation \u2192 more sensitive",
                           rho_mito),
           size = 3.8, color = "grey20") +
  labs(title = "Mitochondrial Translation Signature Predicts Selumetinib Sensitivity",
       subtitle = paste0("Score = mean of: ",
                         paste(paste0(mito_in, " [3s]"), collapse = " + ")),
       x = "Mitochondrial Translation Score", y = "Selumetinib Viability") +
  theme_bw(base_size = 13) +
  theme(legend.position = "right")
print(p_mito)
ggsave("biomarker_figs/final_selumetinib_mito.pdf", p_mito, width = 9, height = 6)
```

---

# 5. Finding 3: Transcriptional CDK Biomarkers (Alvocidib)

## 5.1 Rationale

Alvocidib yielded 4 triple-hit biomarkers: NUDCD3, MRAS, PSMB5, EIF4E. We validated these against
all transcriptional CDK inhibitors (CDK9: Dinaciclib, Enitociclib, Fadraciclib; CDK7: THZ1) and
cell cycle CDK inhibitors (CDK4/6: Palbociclib, Ribociclib, Abemaciclib, Trilaciclib). The key
finding: THZ1 (CDK7) aligns with the CDK9 drugs, not CDK4/6 — confirming these biomarkers reflect
RNA Pol II-dependent transcription rather than the CDK family broadly.

## 5.2 CDK dot plot: transcriptional vs cell cycle specificity

Each panel shows one biomarker; each dot is one drug colored by CDK class. Crossbar would
indicate class mean (omitted for clarity). The pattern is clean: blue dots (CDK9 + CDK7) cluster
high, red dots (CDK4/6) cluster near zero or negative. Alvocidib is highlighted with a black ring
as the discovery drug. MRAS shows the cleanest specificity with all transcriptional CDK drugs at
rho ≥ 0.4 and CDK4/6 drugs near zero.

```{r cdk_dotplot, fig.width=16, fig.height=6}
cdk_targets <- c("NUDCD3", "MRAS", "PSMB5", "EIF4E")
cdk_in <- intersect(cdk_targets, colnames(fm))

cdk_classes <- list(
  "CDK9\n(Transcriptional)" = c("Alvocidib", "Dinaciclib", "Enitociclib", "Fadraciclib"),
  "CDK7\n(Transcriptional)" = c("THZ1"),
  "CDK4/6\n(Cell Cycle)" = c("Palbociclib", "Ribociclib", "Abemaciclib", "Trilaciclib")
)

cdk_cor <- do.call(rbind, lapply(names(cdk_classes), function(cls) {
  drugs <- intersect(cdk_classes[[cls]], colnames(drug_mat_all))
  if (length(drugs) == 0) return(NULL)
  do.call(rbind, lapply(drugs, function(d) {
    do.call(rbind, lapply(cdk_in, function(p) {
      rho <- suppressWarnings(cor(drug_mat_all[shared_val, d], fm[shared_val, p],
                                   method = "spearman", use = "pairwise.complete.obs"))
      data.frame(drug = d, protein = paste0(p, " [3s]"), rho = rho,
                 class = cls, stringsAsFactors = FALSE)
    }))
  }))
}))

cdk_cor$class <- factor(cdk_cor$class, levels = c(
  "CDK9\n(Transcriptional)", "CDK7\n(Transcriptional)", "CDK4/6\n(Cell Cycle)"
))
cdk_cor$protein <- factor(cdk_cor$protein,
                          levels = paste0(c("NUDCD3", "MRAS", "PSMB5", "EIF4E"), " [3s]"))

class_colors_cdk <- c("CDK9\n(Transcriptional)" = "steelblue",
                      "CDK7\n(Transcriptional)" = "lightskyblue",
                      "CDK4/6\n(Cell Cycle)" = "firebrick")

p_dot <- ggplot(cdk_cor, aes(x = class, y = rho, color = class)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.3) +
  geom_point(size = 3.5, alpha = 0.7) +
  geom_point(data = subset(cdk_cor, drug == "Alvocidib"),
             shape = 21, fill = NA, color = "black", stroke = 1, size = 4) +
  geom_text_repel(
    aes(label = drug,
        fontface = ifelse(drug == "Alvocidib", "bold", "plain")),
    size = 3, max.overlaps = 20, segment.color = "grey60", segment.size = 0.3,
    show.legend = FALSE
  ) +
  facet_wrap(~ protein, ncol = 4) +
  scale_color_manual(values = class_colors_cdk) +
  labs(title = "Alvocidib Biomarkers [3s], Transcriptional CDK (blue) vs Cell Cycle CDK (red)",
       x = NULL, y = "Spearman rho with viability") +
  theme_bw(base_size = 13) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 12),
        axis.text.x = element_text(size = 9))

print(p_dot)
ggsave("biomarker_figs/final_cdk_dotplot.pdf", p_dot, width = 16, height = 6)
```

## 5.3 NUDCD3 across the 5 transcriptional CDK inhibitors

Per-drug scatter for NUDCD3 across all 5 transcriptional CDK inhibitors (CDK9 + THZ1). Each panel
shows the rho value; Alvocidib (the discovery drug) is shown first with bold formatting. NUDCD3
abundance correlates positively with viability across all 5 drugs, with rho ranging from 0.41 to
0.73. NUDCD3 is a HSP70/HSP90 co-chaperone — this finding bridges the transcriptional CDK story
to the chaperone proteostasis theme.

```{r nudcd3_transcriptional, fig.width=14, fig.height=5}
cdk_transcriptional <- intersect(
  c("Alvocidib", "Dinaciclib", "Enitociclib", "Fadraciclib", "THZ1"),
  colnames(drug_mat_all))

plots_nudcd3 <- lapply(cdk_transcriptional, function(d) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, d],
    NUDCD3 = fm[shared_val, "NUDCD3"]
  ) %>% filter(is.finite(viability), is.finite(NUDCD3))

  rho <- suppressWarnings(cor(df$NUDCD3, df$viability, method = "spearman"))
  is_disc <- (d == "Alvocidib")
  target_label <- if (d == "THZ1") "(CDK7)" else "(CDK9)"

  ggplot(df, aes(x = NUDCD3, y = viability)) +
    geom_point(aes(color = Patient), size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
             label = sprintf("rho = %.2f", rho), size = 4, color = "grey20") +
    labs(title = if (is_disc) paste0(d, " (CDK9 - discovery)") else paste0(d, " ", target_label),
         x = "NUDCD3 Abundance", y = "Viability") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(face = if (is_disc) "bold" else "plain"))
})

p_nudcd3 <- wrap_plots(plots_nudcd3, ncol = 3) +
  plot_annotation(
    title = "NUDCD3 [3s] Predicts Transcriptional CDK Inhibitor Resistance",
    subtitle = "Triple-hit discovery from Alvocidib. Validated across 4 CDK9 + 1 CDK7 inhibitor."
  )
print(p_nudcd3)
ggsave("biomarker_figs/final_NUDCD3_transcriptional_cdk.pdf", p_nudcd3, width = 14, height = 5)
```

---

# 6. Finding 4: Fimepinostat Dual-Mechanism Amplification

## 6.1 Rationale

Fimepinostat is a dual PI3K/HDAC inhibitor. Its 6 triple-hit biomarkers (RASA4, ETS1, BCL7B,
GALC, ITGA5, LIFR) show correlations 2–4x stronger than any single drug class. We decomposed
the PI3K class into PI3K-selective, PI3K/mTOR dual, HDAC-only, and mTOR-only to identify which
drug arm contributes each biomarker.

## 6.2 Fimepinostat biomarker deconvolution across drug classes

Bar chart per biomarker showing mean rho when computed across each drug class. Fimepinostat's
single-drug rho (red) consistently exceeds the mean rho of any single-class group, confirming that
the dual mechanism amplifies biomarker signal. BCL7B is the only protein where HDAC mean rho
approaches the Fimepinostat value, identifying it as the most HDAC-enriched biomarker.

```{r fime_deconv, fig.width=16, fig.height=8}
pi3k_selective <- intersect(c("Copanlisib", "Alpelisib", "Eganelisib", "Umbralisib", "Duvelisib"),
                            colnames(drug_mat_all))
pi3k_mtor_dual <- intersect(c("Omipalisib", "Apitolisib", "Bimiralisib", "NVP-BEZ235", "LY3023414"),
                            colnames(drug_mat_all))
hdac_only <- intersect(c("Vorinostat", "Panobinostat", "Belinostat", "Romidepsin", "Entinostat"),
                       colnames(drug_mat_all))
mtor_only_drugs <- intersect(c("Rapamycin", "Everolimus", "Temsirolimus",
                                "Sapanisertib", "Vistusertib"), colnames(drug_mat_all))

fime_proteins <- c("RASA4", "ETS1", "BCL7B", "GALC", "ITGA5", "LIFR")
fime_in <- intersect(fime_proteins, colnames(fm))
fime_evidence <- rep(" [3s]", length(fime_proteins)); names(fime_evidence) <- fime_proteins

pi3k_sel_label  <- sprintf("PI3K-selective\n(n=%d)", length(pi3k_selective))
pi3k_dual_label <- sprintf("PI3K/mTOR dual\n(n=%d)", length(pi3k_mtor_dual))
hdac_label      <- sprintf("HDAC only\n(n=%d)", length(hdac_only))
mtor_label      <- sprintf("mTOR only\n(n=%d)", length(mtor_only_drugs))
fime_label      <- "Fimepinostat\n(PI3K+HDAC, n=1)"

class_data <- do.call(rbind, lapply(fime_in, function(p) {
  get_mean <- function(drug_set, cls) {
    if (length(drug_set) == 0) return(NULL)
    rhos <- sapply(drug_set, function(d) {
      suppressWarnings(cor(drug_mat_all[shared_val, d], fm[shared_val, p],
                           method = "spearman", use = "pairwise.complete.obs"))
    })
    data.frame(protein = paste0(p, fime_evidence[p]), class = cls,
               mean_rho = mean(rhos, na.rm = TRUE), stringsAsFactors = FALSE)
  }
  rbind(
    data.frame(protein = paste0(p, fime_evidence[p]), class = fime_label,
               mean_rho = suppressWarnings(cor(drug_mat_all[shared_val, "Fimepinostat"],
                                                fm[shared_val, p], method = "spearman")),
               stringsAsFactors = FALSE),
    get_mean(pi3k_selective, pi3k_sel_label),
    get_mean(pi3k_mtor_dual, pi3k_dual_label),
    get_mean(hdac_only, hdac_label),
    get_mean(mtor_only_drugs, mtor_label)
  )
}))

deconv_palette <- setNames(
  c("firebrick", "steelblue", "slateblue", "darkorange", "grey50"),
  c(fime_label, pi3k_sel_label, pi3k_dual_label, hdac_label, mtor_label)
)
class_data$class <- factor(class_data$class, levels = names(deconv_palette))

p_deconv <- ggplot(class_data, aes(x = class, y = mean_rho, fill = class)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_text(aes(label = sprintf("%.2f", mean_rho)),
            vjust = ifelse(class_data$mean_rho >= 0, -0.3, 1.3), size = 3) +
  facet_wrap(~ protein, ncol = 3, scales = "free_y") +
  scale_fill_manual(values = deconv_palette) +
  labs(title = "Fimepinostat Biomarker Deconvolution",
       subtitle = "Dual PI3K+HDAC mechanism amplifies biomarker signal 2-4x beyond single-class drugs",
       x = NULL, y = "Mean Spearman rho with viability") +
  theme_bw(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        strip.text = element_text(face = "bold"))
print(p_deconv)
ggsave("biomarker_figs/final_fimepinostat_deconvolution.pdf", p_deconv, width = 16, height = 8)
```

## 6.3 RASA4 and ETS1: mirror-image biomarkers for Fimepinostat

RASA4 (RAS GTPase activating protein 4) and ETS1 (ETS proto-oncogene transcription factor) form
a complementary sensitivity/resistance pair with mirrored slopes. RASA4 (rho = -0.68, sensitivity)
inactivates RAS — high RASA4 means tumors are PI3K/mTOR-dependent and collapse when blocked.
ETS1 (rho = +0.72, resistance) is a downstream MAPK transcription factor — high ETS1 means active
MAPK signaling that bypasses PI3K blockade. Together they map the MAPK bypass axis: tumors with
low RASA4 + high ETS1 have constitutive MAPK signaling and resist Fimepinostat.

```{r rasa4_ets1_mirror, fig.width=14, fig.height=5.5}
df_fime <- data.frame(
  sample = shared_val, Patient = patient_val,
  viability = drug_mat_all[shared_val, "Fimepinostat"],
  RASA4 = fm[shared_val, "RASA4"],
  ETS1  = fm[shared_val, "ETS1"]
) %>% filter(is.finite(viability))

rho_rasa4 <- suppressWarnings(cor(df_fime$RASA4, df_fime$viability,
                                   method = "spearman", use = "pairwise.complete.obs"))
rho_ets1  <- suppressWarnings(cor(df_fime$ETS1,  df_fime$viability,
                                   method = "spearman", use = "pairwise.complete.obs"))

p_rasa4_panel <- ggplot(df_fime, aes(x = RASA4, y = viability)) +
  geom_point(aes(color = Patient), size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = pcols) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
           label = sprintf("rho = %.2f (sensitivity)\nHigh RASA4 \u2192 low RAS \u2192 PI3K-dependent",
                           rho_rasa4),
           size = 3.5, color = "grey20") +
  labs(title = "RASA4 [3s] \u2014 Fimepinostat Sensitivity",
       x = "RASA4 Abundance", y = "Viability") +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")

p_ets1_panel <- ggplot(df_fime, aes(x = ETS1, y = viability)) +
  geom_point(aes(color = Patient), size = 3.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = pcols) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
           label = sprintf("rho = %.2f (resistance)\nHigh ETS1 \u2192 active MAPK \u2192 bypasses PI3K block",
                           rho_ets1),
           size = 3.5, color = "grey20") +
  labs(title = "ETS1 [3s] \u2014 Fimepinostat Resistance",
       x = "ETS1 Abundance", y = "Viability") +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")

p_fime_dual <- p_rasa4_panel + p_ets1_panel + plot_layout(guides = "collect") +
  plot_annotation(
    title = "RASA4 and ETS1: Mirror-Image Biomarkers Mapping the MAPK Bypass Axis",
    subtitle = "Both proteins are triple-hit discoveries from Fimepinostat. Opposite slopes reflect opposite biology."
  )
print(p_fime_dual)
ggsave("biomarker_figs/final_fimepinostat_RASA4_ETS1_mirror.pdf", p_fime_dual,
       width = 14, height = 5.5)
```

## 6.4 RASA4 sensitivity across PI3K/mTOR axis

Per-drug scatter showing RASA4 generalizes as a sensitivity marker beyond Fimepinostat across
PI3K-selective, PI3K/mTOR dual, and mTOR drugs. Each panel labels the drug class. RASA4's RAS-GAP
function makes biological sense as a pathway-wide marker — tumors with high RASA4 have suppressed
RAS and depend more on PI3K/mTOR signaling for survival.

```{r rasa4_scatter, fig.width=16, fig.height=8}
rasa4_all_drugs <- intersect(
  c("Fimepinostat", "Omipalisib", "Apitolisib", "Sapanisertib",
    "Copanlisib", "Alpelisib", "Everolimus"),
  colnames(drug_mat_all))

rasa4_class_map <- c(
  "Fimepinostat" = "PI3K+HDAC", "Copanlisib" = "PI3K-selective",
  "Alpelisib" = "PI3K\u03b1-selective", "Omipalisib" = "PI3K/mTOR dual",
  "Apitolisib" = "PI3K/mTOR dual", "Sapanisertib" = "mTORC1/2 catalytic",
  "Everolimus" = "mTORC1 allosteric"
)

plots_rasa4 <- lapply(rasa4_all_drugs, function(d) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, d],
    RASA4 = fm[shared_val, "RASA4"]
  ) %>% filter(is.finite(viability), is.finite(RASA4))

  rho <- suppressWarnings(cor(df$RASA4, df$viability, method = "spearman"))
  cls <- rasa4_class_map[d]
  is_disc <- (d == "Fimepinostat")

  ggplot(df, aes(x = RASA4, y = viability)) +
    geom_point(aes(color = Patient), size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
             label = sprintf("rho = %.2f\n%s", rho, cls), size = 3.3, color = "grey20") +
    labs(title = if (is_disc) paste0(d, " (discovery)") else d,
         x = "RASA4 Abundance", y = "Viability") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(face = if (is_disc) "bold" else "plain"))
})

p_rasa4 <- wrap_plots(plots_rasa4, ncol = 4) +
  plot_annotation(
    title = "RASA4 [3s] Sensitivity Across the PI3K/mTOR Pathway Axis"
  )
print(p_rasa4)
ggsave("biomarker_figs/final_RASA4_classified.pdf", p_rasa4, width = 16, height = 8)
```

## 6.5 BCL7B across HDAC inhibitors

BCL7B (component of the SWI/SNF chromatin remodeling complex) is the most HDAC-enriched biomarker
from the Fimepinostat panel. Per-drug scatter across 6 HDAC inhibitors shows positive rho across
4/5 pure HDAC drugs, validating BCL7B as an HDAC class biomarker rather than a Fimepinostat-only
finding. The chromatin-modifier biology fits — BCL7B-high tumors depend on intact SWI/SNF
function, which is disrupted by HDAC blockade.

```{r bcl7b_hdac, fig.width=14, fig.height=8}
hdac_class_labels <- c(
  "Fimepinostat" = "Fimepinostat\n(PI3K+HDAC)",
  "Panobinostat" = "Panobinostat\n(pan-HDAC)",
  "Romidepsin"   = "Romidepsin\n(HDAC1/2)",
  "Vorinostat"   = "Vorinostat\n(pan-HDAC)",
  "Belinostat"   = "Belinostat\n(pan-HDAC)",
  "Entinostat"   = "Entinostat\n(HDAC1/3)"
)

hdac_for_bcl7b <- intersect(names(hdac_class_labels), colnames(drug_mat_all))

plots_bcl7b <- lapply(hdac_for_bcl7b, function(d) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, d],
    BCL7B = fm[shared_val, "BCL7B"]
  ) %>% filter(is.finite(viability), is.finite(BCL7B))

  rho <- suppressWarnings(cor(df$BCL7B, df$viability, method = "spearman"))

  ggplot(df, aes(x = BCL7B, y = viability)) +
    geom_point(aes(color = Patient), size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
             label = sprintf("rho = %.2f", rho), size = 4, color = "grey20") +
    labs(title = hdac_class_labels[d], x = "BCL7B Abundance", y = "Viability") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none")
})

p_bcl7b <- wrap_plots(plots_bcl7b, ncol = 3) +
  plot_annotation(
    title = "BCL7B [3s] (SWI/SNF Component) Predicts HDAC Inhibitor Resistance",
    subtitle = "Strongest HDAC-enriched biomarker from Fimepinostat deconvolution"
  )
print(p_bcl7b)
ggsave("biomarker_figs/final_BCL7B_hdac.pdf", p_bcl7b, width = 14, height = 8)
```

---

# 7. Finding 5: Copanlisib (PI3K/AKT Bypass)

## 7.1 Rationale

Copanlisib has the lowest variability of all 7 drugs (SD = 0.044), yet yielded 5 triple-hit
biomarkers. NECTIN4 (rho = +0.71) is the strongest: a cell adhesion molecule that directly
activates PI3K/AKT/Rac1 signaling and is the FDA-approved Enfortumab vedotin ADC target.
NECTIN4-high tumors may have constitutive PI3K activation that bypasses pharmacological inhibition.
PCTP validates more broadly across PI3K/mTOR drugs, while C1orf174 provides a sensitivity axis.

## 7.2 NECTIN4 context scatter

Annotated scatter showing the resistance mechanism explicitly. Each sample is labeled, and the
two corner annotations explain the biology: high NECTIN4 → constitutive PI3K/AKT → Copanlisib
bypass; low NECTIN4 → PI3K-dependent → sensitive. NECTIN4 is also the FDA-approved Enfortumab
vedotin target, raising the possibility that NECTIN4-high PI3K-resistant tumors might be ADC
candidates instead.

```{r nectin4_context, fig.width=11, fig.height=8}
df_n4 <- data.frame(
  sample = shared_val, Patient = patient_val,
  viability = drug_mat_all[shared_val, "Copanlisib"],
  NECTIN4 = fm[shared_val, "NECTIN4"]
) %>% filter(is.finite(viability), is.finite(NECTIN4))

rho_n4 <- suppressWarnings(cor(df_n4$NECTIN4, df_n4$viability, method = "spearman"))

p_n4_context <- ggplot(df_n4, aes(x = NECTIN4, y = viability)) +
  geom_point(aes(color = Patient), size = 4.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = pcols) +
  geom_text_repel(aes(label = sample, color = Patient), size = 2.5,
                  max.overlaps = 20, segment.size = 0.3, show.legend = FALSE) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
           label = sprintf("rho = %.2f", rho_n4), size = 5, color = "grey20") +
  annotate("label", x = max(df_n4$NECTIN4, na.rm = TRUE),
           y = min(df_n4$viability, na.rm = TRUE) + 0.02,
           label = "High NECTIN4\n= constitutive PI3K/AKT\n= Copanlisib bypass",
           hjust = 1, size = 3.5, fill = "firebrick", color = "white",
           fontface = "bold", label.size = 0) +
  annotate("label", x = min(df_n4$NECTIN4, na.rm = TRUE),
           y = max(df_n4$viability, na.rm = TRUE) - 0.02,
           label = "Low NECTIN4\n= PI3K-dependent\n= sensitive to Copanlisib",
           hjust = 0, size = 3.5, fill = "steelblue", color = "white",
           fontface = "bold", label.size = 0) +
  labs(title = "NECTIN4 [3s] \u2014 Copanlisib Resistance via PI3K/AKT Bypass",
       subtitle = "NECTIN4 directly activates PI3K/AKT/Rac1. High expression may sustain pathway activation despite PI3K blockade.",
       x = "NECTIN4 Abundance (normalized)", y = "Copanlisib Viability") +
  theme_bw(base_size = 14) +
  theme(legend.position = "right")
print(p_n4_context)
ggsave("biomarker_figs/final_NECTIN4_copanlisib_context.pdf", p_n4_context, width = 11, height = 8)
```

## 7.3 Copanlisib validation heatmap across PI3K/mTOR classes

Heatmap of all 5 Copanlisib triple-hits across 16 PI3K/mTOR drugs, manually grouped by class
(PI3K-selective → PI3K/mTOR dual → PI3K+HDAC → mTOR-only). Copanlisib and Fimepinostat are marked
with asterisks as drugs of interest. Key observations: NECTIN4 is strongest for Copanlisib itself
and Temsirolimus but weaker for other PI3K drugs; PCTP validates more broadly; Fimepinostat shows
near-zero for all 5 proteins despite shared PI3K target.

```{r copa_heatmap, fig.width=12, fig.height=10}
copa_hits <- c("NECTIN4", "PCTP", "C1orf174", "USP38", "REPIN1")
copa_in <- intersect(copa_hits, colnames(fm))

fime_in_copa <- intersect("Fimepinostat", colnames(drug_mat_all))
drug_order_copa <- c(pi3k_selective, pi3k_mtor_dual, fime_in_copa, mtor_only_drugs)

copa_mat <- matrix(NA, nrow = length(drug_order_copa), ncol = length(copa_in),
                   dimnames = list(drug_order_copa, copa_in))
for (d in drug_order_copa) {
  for (p in copa_in) {
    copa_mat[d, p] <- suppressWarnings(
      cor(drug_mat_all[shared_val, d], fm[shared_val, p],
          method = "spearman", use = "pairwise.complete.obs"))
  }
}

row_labels <- rownames(copa_mat)
drug_class_copa <- data.frame(
  Class = case_when(
    row_labels %in% pi3k_selective ~ "PI3K-selective",
    row_labels %in% pi3k_mtor_dual ~ "PI3K/mTOR dual",
    row_labels == "Fimepinostat" ~ "PI3K+HDAC",
    row_labels %in% mtor_only_drugs ~ "mTOR only"
  )
)
rownames(drug_class_copa) <- row_labels

ann_colors_copa <- list(
  Class = c("PI3K-selective" = "steelblue", "PI3K/mTOR dual" = "slateblue",
            "PI3K+HDAC" = "firebrick", "mTOR only" = "grey50")
)

# Mark drugs of interest with asterisk
bold_rows <- row_labels
bold_rows[bold_rows == "Copanlisib"]   <- "Copanlisib*"
bold_rows[bold_rows == "Fimepinostat"] <- "Fimepinostat*"
rownames(copa_mat) <- bold_rows
rownames(drug_class_copa) <- bold_rows

pheatmap::pheatmap(
  copa_mat,
  color = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
  breaks = seq(-1, 1, length.out = 101),
  display_numbers = TRUE, number_format = "%.2f", fontsize_number = 9,
  annotation_row = drug_class_copa, annotation_colors = ann_colors_copa,
  cluster_rows = FALSE, cluster_cols = TRUE,
  gaps_row = c(length(pi3k_selective),
               length(pi3k_selective) + length(pi3k_mtor_dual),
               length(pi3k_selective) + length(pi3k_mtor_dual) + length(fime_in_copa)),
  legend_breaks = c(-1, -0.5, 0, 0.5, 1),
  legend_labels = c("-1", "-0.5", "0", "0.5", "1\n\nSpearman\nrho"),
  main = "Copanlisib [3s] Biomarkers Across PI3K/mTOR Drugs\n* = drug of interest",
  fontsize = 11, fontsize_row = 10, fontsize_col = 10,
  angle_col = 45, cellwidth = 50, cellheight = 30,
  filename = "biomarker_figs/final_copanlisib_validation.pdf",
  width = 12, height = 10
)
```

## 7.4 C1orf174 sensitivity across PI3K/mTOR axis

C1orf174 (MEAK7, recently linked to mTORC1 signaling) consistently predicts sensitivity (negative
rho) across PI3K-selective, PI3K/mTOR dual, and mTOR drugs. Two informative exceptions: Duvelisib
(PI3Kδ/γ-selective) breaks the pattern because it bypasses the mTOR axis, and Fimepinostat is flat
because the HDAC mechanism dominates response biology. The pattern supports C1orf174 as a marker
of PI3Kα→AKT→mTOR-dependent tumors.

```{r c1orf174, fig.width=16, fig.height=8}
c1_drugs <- intersect(c("Copanlisib", "Duvelisib", "Alpelisib", "Apitolisib",
                         "Temsirolimus", "Everolimus", "Omipalisib", "Fimepinostat"),
                      colnames(drug_mat_all))

c1_class_map <- c(
  "Copanlisib" = "PI3K-selective", "Duvelisib" = "PI3K\u03b4/\u03b3-selective",
  "Alpelisib" = "PI3K\u03b1-selective",
  "Apitolisib" = "PI3K/mTOR dual", "Omipalisib" = "PI3K/mTOR dual",
  "Temsirolimus" = "mTOR-selective", "Everolimus" = "mTOR-selective",
  "Fimepinostat" = "PI3K/HDAC dual"
)

plots_c1 <- lapply(c1_drugs, function(d) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, d],
    C1orf174 = fm[shared_val, "C1orf174"]
  ) %>% filter(is.finite(viability), is.finite(C1orf174))

  rho <- suppressWarnings(cor(df$C1orf174, df$viability, method = "spearman"))
  cls <- c1_class_map[d]
  is_disc <- (d == "Copanlisib")

  ggplot(df, aes(x = C1orf174, y = viability)) +
    geom_point(aes(color = Patient), size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
             label = sprintf("rho = %.2f\n%s", rho, cls), size = 3.3, color = "grey20") +
    labs(title = if (is_disc) paste0(d, " (discovery)") else d,
         x = "C1orf174 Abundance", y = "Viability") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(face = if (is_disc) "bold" else "plain"))
})

p_c1 <- wrap_plots(plots_c1, ncol = 4) +
  plot_annotation(
    title = "C1orf174 [3s] \u2014 Sensitivity Marker Across the PI3K/mTOR Axis",
    subtitle = "Consistently negative except Duvelisib (PI3K\u03b4/\u03b3, bypasses mTOR) and Fimepinostat (HDAC-dominant)"
  )
print(p_c1)
ggsave("biomarker_figs/final_C1orf174_sensitivity.pdf", p_c1, width = 16, height = 8)
```

---

# 8. Finding 6: Digoxin (Multi-Factorial Resistance)

## 8.1 Rationale

Digoxin is the mechanistic outlier — a cardiac glycoside targeting Na+/K+-ATPase with no close
replicates in the screen. 6 triple-hit biomarkers were found, each representing a distinct
cellular defense mechanism: drug uptake (OSCP1), electron transfer (CYB561D2), translation
(EEF2), RNA surveillance (ZCCHC8), DNA repair (ABRAXAS1), and chaperone biology (DNAJC8).
DNAJC8 links to the broader HSP chaperone resistance axis.

## 8.2 OSCP1 predicts Digoxin sensitivity

Standalone scatter highlighting the strongest Digoxin biomarker. OSCP1 (Organic Solute Carrier
Partner 1) is a transmembrane transporter; the negative correlation (rho = -0.66) suggests a
pharmacokinetic mechanism — higher OSCP1 → more drug uptake → higher intracellular concentration
→ more cell death. This is the most mechanistically straightforward Digoxin biomarker and would
be testable with transport assays.

```{r oscp1_digoxin, fig.width=9, fig.height=7}
df_oscp1 <- data.frame(
  sample = shared_val, Patient = patient_val,
  viability = drug_mat_all[shared_val, "Digoxin"],
  OSCP1 = fm[shared_val, "OSCP1"]
) %>% filter(is.finite(viability), is.finite(OSCP1))

rho_oscp1 <- suppressWarnings(cor(df_oscp1$OSCP1, df_oscp1$viability, method = "spearman"))

p_oscp1 <- ggplot(df_oscp1, aes(x = OSCP1, y = viability)) +
  geom_point(aes(color = Patient), size = 4, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "grey30",
              linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = pcols) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
           label = sprintf("rho = %.2f (sensitivity)\nHigher OSCP1 \u2192 more drug uptake?",
                           rho_oscp1),
           size = 4, color = "grey20") +
  labs(title = "OSCP1 [3s] Predicts Digoxin Sensitivity",
       subtitle = "Organic solute carrier partner 1. May facilitate digoxin uptake into cells.",
       x = "OSCP1 Abundance (normalized)", y = "Digoxin Viability") +
  theme_bw(base_size = 14) +
  theme(legend.position = "right")
print(p_oscp1)
ggsave("biomarker_figs/final_OSCP1_digoxin.pdf", p_oscp1, width = 9, height = 7)
```

## 8.3 All 6 Digoxin biomarkers

Grid of all 6 Digoxin triple-hits, each panel labeled with the protein's function. The
multi-factorial nature of Digoxin resistance is visible: 1 sensitivity marker (OSCP1, drug uptake)
and 5 resistance markers spanning entirely different cellular processes. This is unique among the
7 drugs — others show thematically coherent biomarkers (chaperone for HSP90, mitochondrial for
MEK), while Digoxin engages diverse defense mechanisms.

```{r digoxin_grid, fig.width=14, fig.height=8}
dig_all <- c("OSCP1", "CYB561D2", "EEF2", "ZCCHC8", "ABRAXAS1", "DNAJC8")
dig_all_in <- intersect(dig_all, colnames(fm))

dig_functions <- c(
  "OSCP1" = "Solute transporter",
  "CYB561D2" = "Electron transfer",
  "EEF2" = "Translation elongation",
  "ZCCHC8" = "RNA surveillance",
  "ABRAXAS1" = "DNA repair (BRCA1)",
  "DNAJC8" = "HSP40 co-chaperone"
)

plots_dig <- lapply(dig_all_in, function(p) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, "Digoxin"],
    abundance = fm[shared_val, p]
  ) %>% filter(is.finite(viability), is.finite(abundance))

  rho <- suppressWarnings(cor(df$abundance, df$viability, method = "spearman"))
  dir <- if (rho < 0) "sensitivity" else "resistance"

  ggplot(df, aes(x = abundance, y = viability)) +
    geom_point(aes(color = Patient), size = 2.5, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
             label = sprintf("rho = %.2f (%s)", rho, dir), size = 3.5, color = "grey20") +
    labs(title = sprintf("%s [3s] \u2014 %s", p, dig_functions[p]),
         x = "Abundance", y = "Viability") +
    theme_bw(base_size = 10) +
    theme(legend.position = "none")
})

p_dig_grid <- wrap_plots(plots_dig, ncol = 3) +
  plot_annotation(
    title = "Digoxin Biomarkers: Multi-Factorial Resistance Profile",
    subtitle = "1 sensitivity marker (OSCP1) + 5 resistance markers spanning translation, RNA surveillance, DNA repair, and chaperone biology. DNAJC8 links to the broader HSP chaperone resistance axis."
  )
print(p_dig_grid)
ggsave("biomarker_figs/final_digoxin_all_hits.pdf", p_dig_grid, width = 14, height = 8)
```

## 8.4 DNAJC8: stronger for Digoxin than for HSP90 inhibitors

DNAJC8 is an HSP40 co-chaperone that appears in the chaperone heatmap, but its strongest signal
is actually for Digoxin (rho = 0.45), not any HSP90 inhibitor (best is SNX-2112 at 0.35). This
suggests DNAJC8's co-chaperone function is more relevant for buffering Digoxin's proteotoxic
effects (ion imbalance from Na+/K+-ATPase disruption causes protein misfolding) than for
buffering pharmacological HSP90 inhibition. DNAJC8 links to the broader HSP chaperone resistance
axis but is primarily a Digoxin biomarker.

```{r dnajc8_comparison, fig.width=14, fig.height=8}
dnajc8_drugs <- intersect(c("Digoxin", "Onalespib", "SNX-2112", "KW-2478",
                             "Ganetespib", "SNX-5422"), colnames(drug_mat_all))

plots_dnajc8 <- lapply(dnajc8_drugs, function(d) {
  df <- data.frame(
    sample = shared_val, Patient = patient_val,
    viability = drug_mat_all[shared_val, d],
    DNAJC8 = fm[shared_val, "DNAJC8"]
  ) %>% filter(is.finite(viability), is.finite(DNAJC8))

  rho <- suppressWarnings(cor(df$DNAJC8, df$viability, method = "spearman"))
  is_dig <- (d == "Digoxin")
  cls <- if (is_dig) "Cardiac glycoside" else "HSP90 inhibitor"

  ggplot(df, aes(x = DNAJC8, y = viability)) +
    geom_point(aes(color = Patient), size = 3, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "grey30",
                linetype = "dashed", linewidth = 0.6) +
    scale_color_manual(values = pcols) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.5,
             label = sprintf("rho = %.2f\n%s", rho, cls), size = 3.3, color = "grey20") +
    labs(title = if (is_dig) paste0(d, " (discovery)") else d,
         x = "DNAJC8 Abundance", y = "Viability") +
    theme_bw(base_size = 11) +
    theme(legend.position = "none",
          plot.title = element_text(face = if (is_dig) "bold" else "plain"))
})

p_dnajc8 <- wrap_plots(plots_dnajc8, ncol = 3) +
  plot_annotation(
    title = "DNAJC8 [3s]: Stronger for Digoxin Than Any HSP90 Inhibitor",
    subtitle = "HSP40 co-chaperone discovered as Digoxin biomarker (rho=0.45). Weaker for HSP90 drugs (rho 0.18-0.35)."
  )
print(p_dnajc8)
ggsave("biomarker_figs/final_DNAJC8_digoxin_vs_hsp90.pdf", p_dnajc8, width = 14, height = 8)
```

---

# 9. Pan-Drug Biomarker Candidates

## 9.1 Rationale

Beyond drug-specific biomarkers, we asked which proteins recur across multiple drugs. Proteins
appearing as top correlates for several of the 7 drugs of interest (independent of mechanism)
may represent fundamental determinants of organoid drug response — proteostasis, mitochondrial
function, or other broad cellular states. We selected the 20 proteins most frequently appearing
in the top 50 correlates across all 7 drugs and visualized their Spearman rho across the panel.

## 9.2 Pan-drug heatmap of recurrent biomarkers

Heatmap of the 20 most recurrent proteins × 7 drugs of interest. Drugs are on rows, proteins on
columns. Clustering reveals two major patterns: (1) a chaperone block (HSP90AA1, HSPA4, CHORDC1)
positively correlated with HSP90 inhibitors and Alvocidib, and (2) a metabolic/mitochondrial
block (NDUFS2, OGDH, UQCRC2) negatively correlated with MEK inhibitors. The separation between
HSP90 and MEK columns highlights how different mechanism classes engage distinct cellular
vulnerabilities — there is no single "pan-drug" biomarker, but there are pan-drug biomarker
themes (proteostasis, metabolism).

```{r pan_drug_heatmap, fig.width=14, fig.height=6}
# Find the 20 most recurrent proteins across the 7 drugs of interest (top-50 correlates per drug)
top50_per_drug <- corr_global$cor_tbl %>%
  filter(drug %in% drugs_matched) %>%
  group_by(drug) %>%
  arrange(desc(abs(cor))) %>%
  slice_head(n = 50) %>%
  ungroup()

recurrent_proteins <- top50_per_drug %>%
  group_by(feature) %>%
  summarise(n_drugs = n_distinct(drug),
            mean_abs_rho = mean(abs(cor), na.rm = TRUE),
            .groups = "drop") %>%
  filter(n_drugs >= 2) %>%
  arrange(desc(n_drugs), desc(mean_abs_rho)) %>%
  slice_head(n = 20)

pan_proteins <- intersect(recurrent_proteins$feature, colnames(fm))

pan_cor_mat <- matrix(NA, nrow = length(drugs_of_interest), ncol = length(pan_proteins),
                       dimnames = list(drugs_of_interest, pan_proteins))
for (d in drugs_of_interest) {
  for (p in pan_proteins) {
    pan_cor_mat[d, p] <- suppressWarnings(
      cor(drug_mat_all[shared_val, d], fm[shared_val, p],
          method = "spearman", use = "pairwise.complete.obs"))
  }
}

pheatmap::pheatmap(
  pan_cor_mat,
  color = colorRampPalette(c("steelblue", "white", "firebrick"))(100),
  breaks = seq(-1, 1, length.out = 101),
  cluster_rows = TRUE, cluster_cols = TRUE,
  display_numbers = TRUE, number_format = "%.2f", fontsize_number = 7,
  legend_breaks = c(-1, -0.5, 0, 0.5, 1),
  legend_labels = c("-1", "-0.5", "0", "0.5", "1\n\nSpearman\nrho"),
  main = "Pan-Drug Biomarker Candidates \u2014 Spearman rho Across All 7 Drugs of Interest",
  fontsize = 10, angle_col = 45,
  filename = "biomarker_figs/final_pan_drug_heatmap.pdf",
  width = 14, height = 6
)
```

---

# 10. Export Tables

```{r export}
write_csv(combined, "biomarker_tables/triple_hit_biomarkers_per_drug.csv")
write_csv(drug_info, "biomarker_tables/drug_of_interest_summary.csv")
write_csv(recurrent, "biomarker_tables/multi_strategy_recurrent.csv")

cat("Exported tables to biomarker_tables/:\n")
cat("  - triple_hit_biomarkers_per_drug.csv\n")
cat("  - drug_of_interest_summary.csv\n")
cat("  - multi_strategy_recurrent.csv\n")
```

---

# Notes

- **Positive rho** = higher protein abundance associated with higher viability (**resistance**)
- **Negative rho** = higher protein abundance associated with lower viability (**sensitivity**)
- All analysis uses **global proteomics only** via `uM_viability` metric
- **[3s]** = triple-hit (passed Spearman correlation + extreme responders + within-patient pairs)
- **[2s]** = 2-strategy recurrent across drugs (passed any 2 strategies, shared across drugs)
- Sample size n = 20 organoids from 9 patients, 6 with multiple tumors
- Feature matrix has 8,248 proteins with complete data across all samples
- Standard FDR correction yields no significant hits; multi-strategy concordance is the recovery approach
- Output figures: `biomarker_figs/final_*.pdf`
- Output tables: `biomarker_tables/*.csv`
